---
title: "Managing and operating objects in R"
author: "Francisco E. Fonturbel"
date: "11/May/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Handling data in R

R is an open source platform that can be used to perform statistical analyses and many other things (genetics, maps, etc.). So, R is an object-based programming language, which means that it is based on creating, manipulating, and operating objects.

R objects are case sensitive, for example, "vector" and "Vector" are considered as different objects.

```{r loading data}
#Let's load the data file
#We will create an object called "data" with this information
data<-read.table("data/cactus.txt", header = T)
attach(data)
```

Let's see if we can see the data file by calling the object:

```{r call}
data
```

OK, it seems to be in order! Now we have to concatenate the three columns into one:

```{r conc}
#Concatenate the data rows of df1 into a single vector r
r<-c(t(as.matrix(data)))
#Assign new variables for the treatment levels and number of observations
f = c("Pop1", "Pop2", "Pop3") #Population names
k = 3 #Number of populations sampled
n = 15 #Number of samples per population
#Create a vector of treatment factors that corresponds to each element of r
tm <- gl(k, 1, n*k, factor(f))
```

In a next step, we need to define the blocks:

```{r blocks}
blk = gl(n, k, k*n)             # blocking factor
```


Now we have our data properly ordered. Let's fit a one-way ANOVA model with this:

```{r ANOVA}
aov.cactus <- aov(r ~ tm + blk)
```

OK, now we have an object called `aov.cactus`, now we'll use the `summary()` function to see the results:

```{r result_aov}
summary(aov.cactus)
```

Our results indicate that cactus height is **significantly different** among the three sampled population. However, the block factor `blk` is not significant, indicating that the influence of environmental heterogeneity within populations is negligible.

## Session

```{r session, echo=TRUE}
sessionInfo()
```