---
title: "01 - Managing and operating objects in R"
author: "Francisco E. Fonturbel"
date: "11/May/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Handling data in R

R is an open source platform that can be used to perform statistical analyses and many other things (genetics, maps, etc.). So, R is an object-based programming language, which means that it is based on creating, manipulating, and operating objects.

R objects are case sensitive, for example, "vector" and "Vector" are considered as different objects. Thus, if you are getting a recurrent error trying to call or operate an object, the first thing to do is check your spelling. Be sure that you are calling your object exactly as it was created (hint: the Environment tab on R Studio, located at the upper right corner, contains a list of all objects and functions currently loaded in R).

## Datasets

The primordial input in science is data. Regardless of our study field, we all use data (collected after hard fieldwork or long lab hours) and the quality of our analyses and papers depend on the quality of our data. Likewise, properly managing our data is crucial to have a smooth experience in R. So, my first advice will be: "please, **AVOID** the `openxlsx` library; it might seem to make your life easier but most of the time it will make your experience with R very hard.

Tips for loading your data without problem:

(1) Prefer csv (comma-separated text) or txt (tab-separated text) formats to input your data in R.

(2) Avoid spaces, points, and commas in variable names.

(3) Do not leave empty cells. If there is no data, use `NA` to fill those empty cells.


### Example datasets

No data to test yet? Don't worry! R has several example datasets to work with. Just type `data()` to see the available data in base R (some additional datasets may be available within other packages as well).

One of the nicests datasets is `iris`, which contains information on flower measurements of three species of the _Iris_ genus.

```{r iris}
data(iris)
attach(iris)
names(iris)
```

Thus, this dataset contains four columns with quantitative data and a fifth column with categorical values, corresponding to the different species.

Using the command `head()` we can examine the first 6 lines of the dataset and figure how it looks like:

```{r irish}
head(iris)
```

How many species are present in this dataset? we may count them one by one or... just ask R:

```{r splev}
#Examining variable levels
levels(Species)

#Creating an object to quantify how many levels do we have
spp<-levels(Species)
length(spp)
```

In this case there are only three species, but this is particularly useful for large datasets with many levels.

## First-aid plots: viewing our data



## Session

```{r session, echo=TRUE}
sessionInfo()
```