---
title: "02 - Making awesome plots in R"
author: "Francisco E. Fonturbel"
date: "24/May/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## The art of making graphics in R

R is a powerful graphic engine too. But we have to learn some things before we can produce awesome plots for our reports and papers. When R become popular, most people acknowledged it statistical power but criticized its limited ability to produce quality graphics. Basic R plots were plain and dull, and often needed tweaking in an external software (e.g., Adobe Illustrator). However, now the situation is quite different thanks to the arrival of `ggplot`, which have introduced a new way to produce high-quality plots of almost everything using an universal grammar.

This lesson contains some examples developed by Naupaka Zimmerman during an ESA workshop in 2019.

```{r packages}
library(ggplot2)
library(ggthemes)
library(tidyr)
library(dplyr)
library(gapminder)
```

So, with the `ggplot2` R library we can go from this:

```{r ugly}
plot(gdpPercap ~ lifeExp, data = gapminder, col = gapminder$continent)
```


To this:

```{r awesome}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(shape = 1, aes(color = continent)) +
  stat_smooth(method = "lm", size = 1, color = "black") +
  scale_x_log10() + 
  xlab("Per Capita GDP") + 
  ylab("Life Expectancy (yrs)") +
  facet_wrap(~continent) +
  theme_few() + 
  guides(color = "none")
```


## Final thoughts

R is a useful graphic tool to produce high-quality plots

## Session

```{r session, echo=TRUE}
sessionInfo()
```