---
title: "03 - Simple hypothesis test in R"
author: "Francisco E. Fonturbel"
date: "01/June/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Simple hypothesis test in R

Now that we have the basic principles of creating and operating objects in R, as well as making some pretty nice graphics, now let's get started with statistics in R. Now we will start with some basic hypothesis testing by comparing the means of two or more treatments (groups).

Let's imagine that we measured the growth of some plants with two treatments: watered with 100 mL per day and watered with 50 mL per day. If there is no effect of the amount of water received, we should expect (i.e., null hypothesis) that the mean height of the plants will be equal:

$\mu {1} = \mu {2}$ 

And, of course, our alternative hypothesis would be:

$\mu {1} \neq \mu {2}$

## Two groups: _t-test_

As we have only two groups, we can use Student's _t test_ using the `t.test` function included in the R base package.

```{r packages}
data<-read.csv("data/03_waterpl.csv", header = TRUE, sep = ";")
attach(data)
```

So, we have a simple dataset containing plant height measurements for two treatments (named A and B).

```{r ttest}
t.test(height~treatment)
```

So, it seems that plant height differs, let's take a look using a simple boxplot:

```{r bp1}
boxplot(height~treatment, col = "darkorange1")
```

Also, _t tests_ depend on two main assumptions: normality and variance homogeneity. To examine normality we will use the Shapiro-Wilk test and to examine variance homogeneity we will use the Bartlett test.

```{r norm}
shapiro.test(height)
hist(height, col = "dodgerblue2")
```

OK, our data is normal but has some deviations from the "ideal" normal curve. Let's examine a Q-Q (quantile-quantile) plot to confirm this:

```{r qq}
qqnorm(data$height, pch = 1, frame = FALSE)
qqline(data$height, col = "firebrick", lwd = 3)
```

An ideal normal distribution will follow the red line.

And now, let's test for **variance homogeneity**

```{r vhom}
bartlett.test(height~treatment)
```

What a relief! our data has homogeneous variances and therefore our results are reliable (failing to meet statistical assumptions would result in unreliable results).

### Non-parametric version

What if we fail to meet the normality assumption? Well, we can try a non-parametric version of the _t test_ called Mann–Whitney–Wilcoxon, which can be summoned using the `wilcox.test` function.

```{r npwil}
wilcox.test(height~treatment)
```

The non-parametric test has confirmed our _t-test_ results.


## More than two groups: the ANOVA test

What if instead two treatments we have three? One first though would be making three _t-tests_ to compare treatment 1 vs 2, treatment 2 vs 3, and treatment 1 vs 3. But the **problem** here is that we have multiple comparisons upon the same data and therefore we will need apply some sort of correction to our significance values (e.g., a Bonferroni correction).

But, we have a more adequate option that is called Analysis of Variance (ANOVA) that can be called using the `aov` function in R. To illustrate this, we will call our old friend, the _iris_ dataset that has three species and some morphological measurements.

```{r iris}
data(iris)
attach(iris)
```

How sepal lenght differ among species? let's make a simple boxplot:

```{r bp_iris}
boxplot(Sepal.Length~Species, col = "blueviolet", ylab= "Sepal lenght")
```

```{r aov}
comp1<-aov(Sepal.Length~Species)
summary(comp1)
```

As we can suspect after examining the boxplot, there are significant differences here. ANOVA tests rely on the same assumptions than _t-tests_ so let's check normality and variance homogeneity:

```{r normsl}
shapiro.test(Sepal.Length)
```

Ouch! our response variable is not normal!


```{r vhiris}
bartlett.test(Sepal.Length~Species)
```

And also variances are not homogeneous. Therefore, our results may not be reliable.


## Final thoughts

a

## Session

```{r session, echo=TRUE}
sessionInfo()
```