---
title: "04 - Other ANOVA designs part 2"
author: "Francisco E. Fonturbel"
date: "14/June/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Completely Randomized Design

Now, we will examine the code and the explanations to perform a **completely randomized design**. In this kind of model, we have only one experimental factor under consideration.

Let's suppose that we sample the height of _Echinopsis chiloensis_ cacti in three populations (randomly selected) out of 15 populations available in our study area. We took 15 measurements at each one of the populations. The information is contained on the file named _04-cactus.txt_ located in the `data` folder.

```{r data}
#Let's load the data file
data<-read.table("data/04-cactus.txt", header = T)
attach(data)
```

Let's see if we can see the data file by calling the object:

```{r call}
data
```

OK, it seems to be in order! Now we have to concatenate the three columns into one:

```{r conc}
#Concatenate the data rows of df1 into a single vector r
r<-c(t(as.matrix(data)))
#Assign new variables for the treatment levels and number of observations
f = c("Pop1", "Pop2", "Pop3") #Population names
k = 3 #Number of populations sampled
n = 15 #Number of samples per population
#Create a vector of treatment factors that corresponds to each element of r
tm <- gl(k, 1, n*k, factor(f))
```

Now we have our data properly ordered. Let's fit a one-way ANOVA model with this:

```{r ANOVA.r}
aov.cactus.r <- aov(r ~ tm)
```

OK, now we have an object called `aov.cactus.r`, now we'll use the `summary()` function to see the results:

```{r result_aov_r}
summary(aov.cactus.r)
```

Our results indicate that cactus height is **significantly different** among the three sampled population.


## Randomized Block Design

Now, let's go further and get acquianted with the **randomized block design**. In this kind of model, similar test subjects are grouped in **blocks** and each block is tested against all treatment levels of the primary factor, at a random order. This approach is intended to get rid of possible influences of external factors (e.g., slope or sunlight incidence).


In a further step, we need to define the blocks:

```{r blocks}
blk = gl(n, k, k*n)             # blocking factor
```


Now we have our data properly ordered blocks, let's fit a one-way ANOVA model with this:

```{r ANOVAb}
aov.cactus.b <- aov(r ~ tm + blk)
```

OK, now we have an object called `aov.cactus`, now we'll use the `summary()` function to see the results:

```{r result_aov_b}
summary(aov.cactus.b)
```

Our results indicate that cactus height is **significantly different** among the three sampled population. However, the block factor `blk` is not significant, indicating that the influence of environmental heterogeneity within populations is negligible.


## Final thoughts

ANOVA is quite a flexible test that allows to consider some sources of variation to obtain more reliable results. If we suspect that external factors may be affecting our data, using a randomized block design may be a good option.

## Session

```{r session, echo=TRUE}
sessionInfo()
```