---
title: "05 - Other ANOVA designs part 2"
author: "Francisco E. Fonturbel"
date: "14/June/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Nested ANOVA

In some opportunities, we have more than one factor explaining our response variable. If those factors are independent, we simply use a two-way ANOVA design, but what if one factor is a subset of the other? Let's imagine that we measured plant diameter in three populations, and at each population we set three experimental treatments: control, half water and 10% water. So, we will have the water treatment factor nested within the population factor. Thus, each treatment constitutes a sub-replicate of each population replicate. Therefore, it won't be correct to treat this as a two-way ANOVA; we need to use a Nested ANOVA design instead. Let's see some example data:

```{r data}
#Let's load the data file
data<-read.table("data/05-infest.txt", header = T)
attach(data)

data$valley <- factor(data$valley, ordered=T)
data$site <-factor(data$site, ordered=T)
```

```{r data_pres}
data
```

As you may see in the example data, we have two valleys (Elqui and Limari) and within each valley we have two sampling sites in which we measured cockroach infestation (ewww!). So, we need to nest `site` within `valley` to properly analyze our data.

Let's fit a nested ANOVA model with this:

```{r ANOVA.r}
replications(~ valley * site, data)

aov.infest <- aov(infestation ~ valley/site)
```

OK, now we have an object called `aov.infest`, now we'll use the `summary()` function to see the results:

```{r result_aov_r}
summary(aov.infest)
```

Our results indicate that infestation height is **significantly different** among valleys and sampling sites.

Another thought... what if we want to know which sites explain the differences? Let me introduce you the _a posteriori_ tests, which are commonly used to explain pairwise differences in ANOVA models. One of the most famous _a posteriori_ tests is the Tukey test, which can be summoned using the `TukeyHSD` function. Let's see:

```{r tukey}
TukeyHSD(aov.infest)
```

Well, some comparisons appear marked as "NA" (which represents **missing values** in R jargon). Why there are missing comparisons? because we have nested factors! of course! we cannot compare sites that are in other valleys.


## Factorial ANOVA

Now, let's go further and get acquianted with the **randomized block design**. In this kind of model, similar test subjects are grouped in **blocks** and each block is tested against all treatment levels of the primary factor, at a random order. This approach is intended to get rid of possible influences of external factors (e.g., slope or sunlight incidence).


In a further step, we need to define the blocks:

```{r blocks}
blk = gl(n, k, k*n)             # blocking factor
```


Now we have our data properly ordered blocks, let's fit a one-way ANOVA model with this:

```{r ANOVAb}
aov.cactus.b <- aov(r ~ tm + blk)
```

OK, now we have an object called `aov.cactus`, now we'll use the `summary()` function to see the results:

```{r result_aov_b}
summary(aov.cactus.b)
```

Our results indicate that cactus height is **significantly different** among the three sampled population. However, the block factor `blk` is not significant, indicating that the influence of environmental heterogeneity within populations is negligible.


## Final thoughts

ANOVA is quite a flexible test that allows to consider some sources of variation to obtain more reliable results. If we suspect that external factors may be affecting our data, using a randomized block design may be a good option.

## Session

```{r session, echo=TRUE}
sessionInfo()
```