---
title: "05 - Other ANOVA designs part 2"
author: "Francisco E. Fonturbel"
date: "28/September/2025"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(ggplot2)
library(dplyr)
library(car) # For Type II/III ANOVA
```

## Nested ANOVA

In some opportunities, we have more than one factor explaining our response variable. If those factors are independent, we simply use a two-way ANOVA design, but what if one factor is a subset of the other? Let's imagine that we measured plant diameter in three populations, and at each population we set three experimental treatments: control, half water and 10% water. So, we will have the water treatment factor nested within the population factor. Thus, each treatment constitutes a sub-replicate of each population replicate. Therefore, it won't be correct to treat this as a two-way ANOVA; we need to use a Nested ANOVA design instead.

**Key concept:** In nested designs, the levels of one factor (e.g., sites) are unique within each level of another factor (e.g., valleys). This is different from crossed designs where all combinations of factors are possible.

Let's see some example data:

```{r data}
#Let's load the data file
data<-read.table("data/05-infest.txt", header = T)
attach(data)

data$valley <- factor(data$valley, ordered=T)
data$site <-factor(data$site, ordered=T)
```

```{r data_pres}
data
```

As you may see in the example data, we have two valleys (Elqui and Limari) and within each valley we have two sampling sites in which we measured cockroach infestation (ewww!). So, we need to nest `site` within `valley` to properly analyze our data.

Let's first check our experimental design balance:

```{r design_check}
# Check the structure of our nested design
table(data$valley, data$site)

# Summary statistics by groups
data %>% 
  group_by(valley, site) %>% 
  summarise(n = n(), 
            mean_infest = mean(infestation),
            sd_infest = sd(infestation),
            .groups = 'drop')
```

Now let's fit a nested ANOVA model:

```{r ANOVA.r}
replications(~ valley * site, data)

aov.infest <- aov(infestation ~ valley/site)
```

OK, now we have an object called `aov.infest`, now we'll use the `summary()` function to see the results:

```{r result_aov_r}
summary(aov.infest)
```

Our results indicate that infestation height is **significantly different** among valleys and sampling sites within valleys. Notice how the F-test for sites uses the residual error as denominator, which is appropriate for this nested design.

Another thought... what if we want to know which sites explain the differences? Let me introduce you the _a posteriori_ tests, which are commonly used to explain pairwise differences in ANOVA models. One of the most famous _a posteriori_ tests is the Tukey test, which can be summoned using the `TukeyHSD` function. Let's see:

```{r tukey}
TukeyHSD(aov.infest)
```

Well, some comparisons appear marked as "NA" (which represents **missing values** in R jargon). Why there are missing comparisons? because we have nested factors! of course! we cannot compare sites that are in other valleys - Site 1 in Elqui valley is completely different from Site 1 in Limari valley.

```{r nested_visualization}
# Let's visualize our nested design
boxplot(infestation ~ valley:site, 
        data = data, 
        col = c("lightblue", "lightgreen", "salmon", "wheat"),
        main = "Cockroach Infestation by Valley and Site",
        xlab = "Valley:Site", 
        ylab = "Infestation Level")
```

## Factorial ANOVA

Until now we have operated with **fixed factors**. A fixed factor is a categorical variable from which we have total control and we have measured all its levels of interest. But not everything can be labeled as a fixed factor in nature. We may have other confounding factors that can be affecting our results, but we have no control over them and/or we have (or cannot) measured all levels. Those factors are called **random factors**.

### Understanding Factor Types

**Fixed factors:** 
- Levels are specifically chosen by the researcher
- Inference applies only to the tested levels
- Examples: specific treatments, particular species, chosen temperatures

**Random factors:**
- Levels are randomly sampled from a larger population
- Inference extends to the entire population of levels
- Examples: randomly selected individuals, random sampling locations, random time periods

### Factorial ANOVA Models

In Factorial ANOVA designs we have three possible model types:

1. **Model I (Fixed Effects Model):** Have only fixed factors.
2. **Model II (Random Effects Model):** Have only random factors.
3. **Model III (Mixed Effects Model):** Have a mixture of fixed and random factors.

Let's use an example from Quinn (1988), in which we have the mean number of eggs as a response variable with two factors: season (fixed factor) and density treatments (fixed factor).

```{r data2}
dataf<-read.csv("data/05-quinn.csv", header = TRUE, sep = ",")
attach(dataf)

# Convert to factors and check structure
dataf$SEASON <- factor(dataf$SEASON)
dataf$DENSITY <- factor(dataf$DENSITY)

str(dataf)
head(dataf)
```

### Exploratory Data Analysis

Before diving into the analysis, let's explore our data:

```{r exploration}
# Summary statistics
dataf %>% 
  group_by(SEASON, DENSITY) %>% 
  summarise(n = n(),
            mean_eggs = mean(EGGS),
            sd_eggs = sd(EGGS),
            se_eggs = sd(EGGS)/sqrt(n()),
            .groups = 'drop')

# Check for balance in our design
table(dataf$SEASON, dataf$DENSITY)
```

### Step-by-step Analysis

Let's start by fitting individual models for each factor to understand their independent effects:

```{r aov1}
aov.season<-aov(EGGS~SEASON, data = dataf)
summary(aov.season)
```

Season has a significant effect on egg number (p < 0.05).

Now let's examine density effects:

```{r aov2}
aov.dens<-aov(EGGS~DENSITY, data = dataf)
summary(aov.dens)
```

Density also has a significant effect. However, examining factors separately doesn't tell us about their joint effects or potential interactions.

### Two-way ANOVA: Additive Model

Let's conduct a two-way ANOVA with additive effects (no interaction):

```{r aov3}
aov.2w<-aov(EGGS~SEASON+DENSITY, data = dataf)
summary(aov.2w)
```

Both factors remain significant when considered together. The additive model assumes that the effect of one factor is consistent across all levels of the other factor.

### Two-way ANOVA: Interaction Model

Now let's examine the interaction model (the most complete model):

```{r aov4}
aov.int<-aov(EGGS~SEASON*DENSITY, data = dataf)
summary(aov.int)
```

**Interpretation:** 
- Both main effects (SEASON and DENSITY) are significant
- The interaction term (SEASON:DENSITY) is not significant (p > 0.05)
- This means the effect of season is consistent across density levels, and vice versa

A **significant interaction** would indicate that the effect of one factor depends on the level of the other factor - for example, if season effects were stronger at high densities than at low densities.

### Interaction-only Model

If we want to test ONLY the interaction term:

```{r aov5}
aov.intonly<-aov(EGGS~SEASON:DENSITY, data = dataf)
summary(aov.intonly)
```

This model tests whether the combination of season and density levels produces effects beyond what would be expected from their individual contributions.

### Visualization and Interpretation

Let's create better visualizations to understand our results:

```{r visualization, fig.width=12, fig.height=8}
# Create interaction plot
par(mfrow=c(2,2))

# Boxplot by combination
boxplot(EGGS~SEASON:DENSITY, 
        col = c("lightblue", "lightgreen", "salmon", "wheat"),
        data = dataf,
        main = "Eggs by Season and Density Combination",
        xlab = "Season:Density",
        ylab = "Number of Eggs")

# Interaction plot - Season as factor
interaction.plot(dataf$DENSITY, dataf$SEASON, dataf$EGGS,
                main = "Interaction Plot: Season by Density",
                xlab = "Density", ylab = "Mean Eggs",
                col = c("blue", "red"), lwd = 2)

# Interaction plot - Density as factor  
interaction.plot(dataf$SEASON, dataf$DENSITY, dataf$EGGS,
                main = "Interaction Plot: Density by Season", 
                xlab = "Season", ylab = "Mean Eggs",
                col = c("darkgreen", "orange", "purple"), lwd = 2)

# Boxplot by individual factors
boxplot(EGGS~SEASON, data = dataf, col = "lightblue",
        main = "Eggs by Season", ylab = "Number of Eggs")
```

**Interpreting interaction plots:**
- Parallel lines = no interaction
- Non-parallel lines = interaction present
- Crossing lines = strong interaction

### Model Comparison and Selection

Let's compare our models using AIC (Akaike Information Criterion):

```{r model_comparison}
# Compare models
models <- list(
  "Season only" = aov.season,
  "Density only" = aov.dens, 
  "Additive" = aov.2w,
  "Interaction" = aov.int
)

# Calculate AIC for each model
aic_values <- sapply(models, AIC)
aic_table <- data.frame(
  Model = names(aic_values),
  AIC = aic_values,
  Delta_AIC = aic_values - min(aic_values)
)
aic_table[order(aic_table$AIC),]
```

The model with the lowest AIC is generally preferred, with differences > 2 being meaningful.

### Post-hoc Tests

Since we have significant main effects, let's examine pairwise differences:

```{r posthoc}
# Tukey HSD for the interaction model
TukeyHSD(aov.int)

# For a more detailed view of main effects
TukeyHSD(aov.int, which = "SEASON")
TukeyHSD(aov.int, which = "DENSITY")
```

### Model Assumptions

Let's check our model assumptions:

```{r assumptions, fig.width=10, fig.height=8}
par(mfrow=c(2,2))
plot(aov.int)

# Additional tests
# Homogeneity of variances (Levene's test)
leveneTest(EGGS ~ SEASON * DENSITY, data = dataf)

# Normality of residuals (Shapiro-Wilk test)
shapiro.test(residuals(aov.int))
```

### Effect Sizes

Understanding statistical significance is important, but effect sizes help us understand practical significance:

```{r effect_sizes}
# Calculate eta-squared (proportion of variance explained)
ss_table <- summary(aov.int)[[1]]
ss_total <- sum(ss_table[,"Sum Sq"])

eta_squared <- ss_table[1:3,"Sum Sq"] / ss_total
names(eta_squared) <- rownames(ss_table)[1:3]

print("Eta-squared (effect sizes):")
eta_squared
```

**Effect size interpretation:**
- Small: η² ≈ 0.01
- Medium: η² ≈ 0.06  
- Large: η² ≈ 0.14

### Including Error Terms (Mixed Effects)

In some cases, we might want to treat one factor as random. Let's assume DENSITY represents randomly selected density levels from a broader population:

```{r mixed_model}
# Treating DENSITY as a random factor
aov.mixed <- aov(EGGS ~ SEASON + Error(DENSITY), data = dataf)
summary(aov.mixed)
```

**Interpretation:** This analysis examines the effects of season on egg number, while accounting for the random variation among density levels. This approach is useful when we want to generalize beyond the specific density levels tested.

### Advanced: Using Type II/III ANOVA

For unbalanced designs or when order matters, we might want Type II or III ANOVA:

```{r type_anova}
# Type II ANOVA (recommended for balanced designs without interactions)
Anova(aov.int, type = "II")

# Type III ANOVA (for unbalanced designs or when interactions are present)
Anova(aov.int, type = "III")
```

## Final thoughts

Nested and Factorial ANOVA designs allow us to analyze more complex experiments and obtain reliable estimates taking multiple sources of variation into consideration. Key considerations include:

1. **Design planning:** Think about your factors (fixed vs. random) and their relationships (nested vs. crossed) before data collection
2. **Model selection:** Use appropriate models based on your experimental design and research questions
3. **Assumption checking:** Always verify that your data meets ANOVA assumptions
4. **Effect sizes:** Don't rely solely on p-values; consider practical significance
5. **Visualization:** Plots help interpret complex interactions and communicate results effectively

Remember that if you perform a flawed design and collect field data, analyzing such data will be difficult and your results might be unreliable or not generalizable to your intended population.

## Session

```{r session, echo=TRUE}
sessionInfo()
```