---
title: "06 - Other ANOVA designs part 3"
author: "Francisco E. Fonturbel"
date: "28/June/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Repeated measures ANOVA

A common task is measuring the same experimental unit in multiple times. For example, we have plants growing under three different treatments and we measure plant height as a response variable, but we measure our plant every week for two months. So, we cannot consider each measurement as an independent data point as they represent pseudo-replicates. Thus, we cannot run a simple one-way ANOVA with such data, but we can run a **repeated measures ANOVA** instead to correct our data indicating R that we have sampled the same experimental unit _t_ times.

```{r data}
#Let's load the data file
data<-read.table("data/06-repeated.txt", header = T)

data$id <- factor(data$id)
data$workout <- factor(data$workout)
data$diet <- factor(data$diet)
data$time <- factor(data$time)
attach(data)
```


Our example data is based on some athletes that had three different workout schemes and three different diets, and we measured their pulse (as a response variable) three times during the experiment. Thus, our fixed factors are workout scheme and diet (athletes received different combinations of those factors), and the column _time_ represent the repeated measured.


Let's fit a repeated measures ANOVA model with this:

```{r ANOVA.r}
rm.aov <- aov(pulse ~ diet * workout * time + Error(id) )
```

OK, now we have an object called `rm.aov`, now we'll use the `summary()` function to see the results:

```{r result_aov_r}
summary(rm.aov)
```

Our show that diet had a **significantly effect** on athletes' pulse, but not workout scheme or the intearaction of both. Further, the repeated measures showed that pulse changed over time and this change was related to diet.


## ANCOVA (Analysis of Covariance)

So far we have used categorical factors, but what about continuous covariates? We may have a non-categorical variable that is not our analysis focus we know that it can influence our results (e.g., altitute). Therefore, in these cases we need to use an **Analysis of Covariance** (ANCOVA for short).

Let's use an example data in which we have fecundity as a response variable, site and microhabitat as fixed factors, and individual size as a continuous covariate.

```{r data2}
library(car)

dataf<-read.table("data/06-ancova.txt", header = TRUE)

dataf$site <- factor(dataf$site, ordered=T)
dataf$microhab <-factor(dataf$microhab, ordered=T)

attach(dataf)
```


And now we can fit the `ANCOVA`:

```{r ancova}
fecund.lm <- lm(lnfecund ~ size + site * microhab)
fecund.ancova <- Anova(fecund.lm, type = 'III')
summary(fecund.lm)
print(fecund.ancova)
```

Our results show that our covariate has a significant effect on our results altogether with site and microhabitat.

Now let's examine some plots to see how residuals are distributed:

```{r acvplot}
par(mfrow=c(2,2))
plot(fecund.lm)
par(mfrow=c(1,1))
```


## Final thoughts

Repeated measures ANOVA and ANCOVA allow us to conduct more complex ANOVA models by including temporal replication (and overcome the pseudoreplication problem) and to include a continuous covariate to account to such variability in our comparisons.

## Session

```{r session, echo=TRUE}
sessionInfo()
```