---
title: "08 - Linear regression and correlation"
author: "Francisco E. Fonturbel"
date: "02/August/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Linear regression

So far, we have dealt with a continuous response variable contrasted against categorical predictors. But in many cases we have continuous predictors. For example, we may assess the effect of temperature on enzymatic activity based on an experimental trial. We know from the books that many enzymes perform better at certain temperatures, and we want to test it in our laboratory. So, let's load the example file with temperature and enzymatic activity data.

```{r data}
data<-read.csv("data/08_enzactivity.csv", header = T, sep = ",")
attach(data)
head(data)
```

Before fitting a linear model, we will examine the distribution of the response variable (i.e., enzymatic activity) to assess if it follows a normal distribution:

```{r norm}
hist(enzactiv, col = "dodgerblue1")
shapiro.test(enzactiv)
```

Well, it is more or less normal...


### Linear relationships

Linear models are very common in science, as many of the relationships that we assume that variables increase or decrease along other predictors, resembling a linear function _y = ax + b_

Let's take a look of our data:

```{r plot1}
plot(enzactiv~temperature, col = "firebrick2", xlab = "Temperature (ºC)", ylab = "Enzymatic activity")
```

Well, it clearly resembles a linear relationship.


## Final thoughts

aaa


## Session

```{r session, echo=TRUE}
sessionInfo()
```