---
title: "08 - Linear regression and correlation"
author: "Francisco E. Fonturbel"
date: "02/August/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Linear regression

So far, we have dealt with a continuous response variable contrasted against categorical predictors. But in many cases we have continuous predictors. For example, we may assess the effect of temperature on enzymatic activity based on an experimental trial. We know from the books that many enzymes perform better at certain temperatures, and we want to test it in our laboratory. So, let's load the example file with temperature and enzymatic activity data.

```{r data}
data<-read.csv("data/08_enzactivity.csv", header = T, sep = ",")
attach(data)
head(data)
```

Before fitting a linear model, we will examine the distribution of the response variable (i.e., enzymatic activity) to assess if it follows a normal distribution:

```{r norm}
hist(enzactiv, col = "dodgerblue1")
shapiro.test(enzactiv)
```

Well, it is more or less normal...


### Linear relationships

Linear

## Effects of the animal tissue

In this first part we will use metal concentration on the animal tissue as our response variable. For this, we must create a response vector first:

```{r y_vector}
y<-data[10:16]
```

Now, we will use the `manova` function and operate this just like a one-way ANOVA:

```{r manova0}
manova0<-manova(as.matrix(y)~Site)
summary(manova0)

boxplot(as.matrix(y)~Site, col ="darkorange2")
```

According to this result, the heavy metal loads are significantly different between polluted and non-polluted sites. While ANOVA reports the _F_ statistic, MANOVA reports Pillai trace instead. However, there are other statistics that can be used with ANOVA, we just need to specify them in the code:

```{r fstats}
summary(manova0, test="Wilks")
summary(manova0, test="Roy")
summary(manova0, test="Hotelling-Lawley")
```

The results are exactly the same, but we have different statistics in the report.

Wait... what if we want to know which metals have different between sites? Well, this can be easily answered by using the function `summary.aov` instead:

```{r aovs}
summary.aov(manova0)
```

Awesome! with this code we can determine that Cd and Zu are not different between sites, but the other metals present significant differences.

### Assumptions

As happens with ANOVA, MANOVA relies on some assumptions. Let's take a look to these plots first:

```{r nplot1}
par(mfcol=c(2,2))
hist(residuals(manova0))
plot(manova0$residuals~manova0$fitted.values)
qqnorm(manova0$residuals)
qqline(manova0$residuals)
par(mfcol=c(1,1))
```

Uhmm... it seems that our data is not normal. Let's use the `mvn` function from the `MVN` package to test it:

```{r mvn1}
multinorm<-mvn(y, mvnTest = "royston", univariatePlot = "qqplot")
multinorm
```

So, we have some normality issues. As an alternative, there is a non-parametric version of MANOVA. We will use the function `adonis` from the `vegan` package to do this:

```{r m0np}
manova0np<-adonis(as.matrix(y)~Site)
manova0np
```

The non-parametric test confirmed our results.

### Testing the other factors

Now that we know that factor _Site_ is important, let's examine the other factors:

**Contact with polluted plants**

```{r manova1}
manova1<-manova(as.matrix(y)~Contact)
summary(manova1)
```

No effect of contact.


**Eating polluted plants**

```{r manova2}
manova2<-manova(as.matrix(y)~Eat)
summary(manova2)
```

No effect as well.


**Males versus females**

```{r manova3}
manova3<-manova(as.matrix(y)~Sex)
summary(manova3)
```

And now we have a significant effect of animal sex on metal loads. Let's make a boxplot:

```{r bps}
boxplot(as.matrix(y)~Sex, col = "dodgerblue1")
```


### Interactions

As with ANOVA, we can have more than one factor in the model and estimate their interactions:

```{r manova4}
manova4<-manova(as.matrix(y)~Site*Sex)
summary(manova4)
```


## Effects on the environment

Now, we will repeat the analyses using the data from soil samples instead.

```{r vector-d}
d<-data[3:9]
```

Now that we have our new response vector, we repeat the site MANOVA:

```{r md}
manovad<-manova(as.matrix(d)~Site)
summary(manovad)

boxplot(as.matrix(d)~Site, col = "firebrick1")
```

We can also examine the individual responses:

```{r aovd}
summary.aov(manovad)
```

In this case, all elements have significantly different concentrations in the soil between sites.

### Testing the other factors


**Contact with polluted plants**

```{r manova5}
manova5<-manova(as.matrix(d)~Contact)
summary(manova5)
```

No effect of contact.


**Eating polluted plants**

```{r manova6}
manova6<-manova(as.matrix(d)~Eat)
summary(manova6)
```

No effect as well.


**Males versus females**

```{r manova7}
manova7<-manova(as.matrix(d)~Sex)
summary(manova7)
```

And no effect as well.


## Final thoughts

MANOVA is a very good alternative when we have multiple response variables and we want to assess the overall effect. In operational terms, MANOVA is operated pretty much like ANOVA. The only missing thing here are the _a posteriori_ tests, as they are not available (yet) for MANOVA.

## Session

```{r session, echo=TRUE}
sessionInfo()
```