---
title: "09 - Multiple linear regression"
author: "Francisco E. Fonturbel"
date: "07/November/2025"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Multiple linear regression

Now that we know how to fit linear models, let's try something bolder. Why limiting us to a single predictor variable? In biological sciences, many phenomena depend on two or more variables than can be used to explain our response variable. Thus, the multivariate extension or linear regression is called **Multiple linear regression**. In this case, our linear model looks like this:

_y = a + b1x1 + b2x2 + b3x3 + ... + bnxn_

Being y the response variable, a and b's the coefficients, and x's the predictor variables in the model.

So, we will use the same `lm()` function but incorporating more than one predictor variable in the formula.

Let's use the `mtcars` dataset (included in R) to make an example of this

```{r data}
library(confintr)

data<-mtcars[,c("mpg","disp","hp","wt")]
attach(data)
head(data)
```

So, we have information on several car brands and models, obtained from their technical specifications. In this case, we want to know which factors determine car's efficiency in fuel use (represented by the variable mpg = miles per gallon) regarding three factors: cylinder displacement (disp), power (hp), and car's weight (wt). Let's fit a linear model for this:

```{r lm1}
mod1<-lm(mpg~disp+hp+wt, data = data)
summary(mod1)
```

As we can see here, cylinder displacement has no significant effect on gas consumption, but power and weight do have a significant effect, with negative relationships (i.e., the more powerful and heavy vehicle, the less gas efficiency).

So, now that we now which variables are important, we may fit a second model examining the interaction between these variables:

```{r lm2}
mod2<-lm(mpg~hp*wt, data = data)
summary(mod2)
```

Great! We now know that there is also a significant interaction between power and weight, so powerful and heavy vehicles will demand a lot of money from out pockets when filling the tank.

### How good is our model?

As in univariate linear regression, we may examine R-squared and residual distribution to determine how well our model describes the data.

**R-squared**

For the first model: 
```{r rsq1}
summary(mod1)$r.squared
```

For the second model: 
```{r rsq2}
summary(mod2)$r.squared
```

(not bad!)

**Residuals**

```{r resids1}
plot(density(resid(mod1))) #A density plot
qqnorm(resid(mod1)) # A quantile normal plot - good for checking normality
qqline(resid(mod1))
```

Uhmmm.... are those residuals normal?

```{r normres1}
shapiro.test(resid(mod1))
```

No, they are not. Let's take a look to the second model instead:

```{r resids2}
plot(density(resid(mod2))) #A density plot
qqnorm(resid(mod2)) # A quantile normal plot - good for checking normality
qqline(resid(mod2))
```

Better, but are those residuals normal?

```{r normres2}
shapiro.test(resid(mod2))
```

Great, now they are normal.

Last but not least, let's make a plot!

```{r mod2_plot}
library(ggplot2)
library(dplyr)
library(gridExtra)

# Assuming mod2 is already fitted: mod2 <- lm(mpg ~ hp * wt, data = mtcars)

# 1. 3D-style interaction plot showing the fitted surface
# Create a grid of hp and wt values for prediction
hp_seq <- seq(min(data$hp), max(data$hp), length.out = 30)
wt_seq <- seq(min(data$wt), max(data$wt), length.out = 30)
pred_grid <- expand.grid(hp = hp_seq, wt = wt_seq)
pred_grid$mpg_pred <- predict(mod2, newdata = pred_grid)

# Plot 1: Contour plot showing the interaction surface
p1 <- ggplot() +
  geom_contour_filled(data = pred_grid, aes(x = hp, y = wt, z = mpg_pred), alpha = 0.7) +
  geom_point(data = data, aes(x = hp, y = wt, size = mpg), shape = 21, 
             fill = "white", color = "black", stroke = 1.5) +
  scale_fill_viridis_d(name = "Predicted\nMPG") +
  labs(title = "MPG as a Function of Horsepower and Weight",
       subtitle = "Interaction model: mpg ~ hp * wt",
       x = "Horsepower (hp)", 
       y = "Weight (1000 lbs)",
       size = "Observed\nMPG") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "right")

# Plot 2: Interaction plot - effect of hp at different wt levels
wt_levels <- quantile(data$wt, probs = c(0.25, 0.5, 0.75))
pred_data_wt <- expand.grid(
  hp = seq(min(data$hp), max(data$hp), length.out = 100),
  wt = wt_levels
)
pred_data_wt$mpg_pred <- predict(mod2, newdata = pred_data_wt)
pred_data_wt$wt_category <- factor(pred_data_wt$wt,
                                    labels = c("Light (25th %ile)", 
                                             "Medium (50th %ile)", 
                                             "Heavy (75th %ile)"))

p2 <- ggplot(pred_data_wt, aes(x = hp, y = mpg_pred, color = wt_category, 
                                linetype = wt_category)) +
  geom_line(size = 1.2) +
  geom_point(data = data, aes(x = hp, y = mpg), 
             color = "black", alpha = 0.5, inherit.aes = FALSE) +
  scale_color_manual(values = c("#2E86AB", "#A23B72", "#F18F01")) +
  labs(title = "Effect of Horsepower on MPG at Different Weight Levels",
       subtitle = "Lines show predicted values from interaction model",
       x = "Horsepower (hp)",
       y = "Miles per Gallon (MPG)",
       color = "Vehicle Weight",
       linetype = "Vehicle Weight") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom")

# Plot 3: Interaction plot - effect of wt at different hp levels  
hp_levels <- quantile(data$hp, probs = c(0.25, 0.5, 0.75))
pred_data_hp <- expand.grid(
  wt = seq(min(data$wt), max(data$wt), length.out = 100),
  hp = hp_levels
)
pred_data_hp$mpg_pred <- predict(mod2, newdata = pred_data_hp)
pred_data_hp$hp_category <- factor(pred_data_hp$hp,
                                    labels = c("Low Power (25th %ile)", 
                                             "Medium Power (50th %ile)", 
                                             "High Power (75th %ile)"))

p3 <- ggplot(pred_data_hp, aes(x = wt, y = mpg_pred, color = hp_category, 
                                linetype = hp_category)) +
  geom_line(size = 1.2) +
  geom_point(data = data, aes(x = wt, y = mpg), 
             color = "black", alpha = 0.5, inherit.aes = FALSE) +
  scale_color_manual(values = c("#06A77D", "#D4A017", "#C73E1D")) +
  labs(title = "Effect of Weight on MPG at Different Horsepower Levels",
       subtitle = "Lines show predicted values from interaction model",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon (MPG)",
       color = "Engine Power",
       linetype = "Engine Power") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom")

# Plot 4: Observed vs Predicted values
data_with_pred <- data
data_with_pred$predicted <- fitted(mod2)
data_with_pred$residuals <- residuals(mod2)

p4 <- ggplot(data_with_pred, aes(x = predicted, y = mpg)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", 
              color = "red", size = 1) +
  geom_point(size = 3, alpha = 0.7, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue", 
              fill = "lightblue", alpha = 0.3) +
  labs(title = "Observed vs Predicted MPG",
       subtitle = paste0("R² = ", round(summary(mod2)$r.squared, 3)),
       x = "Predicted MPG",
       y = "Observed MPG") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))

# Plot 5: Residuals vs Fitted
p5 <- ggplot(data_with_pred, aes(x = predicted, y = residuals)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", size = 1) +
  geom_point(size = 3, alpha = 0.7, color = "steelblue") +
  geom_smooth(se = TRUE, color = "darkblue", fill = "lightblue", alpha = 0.3) +
  labs(title = "Residuals vs Fitted Values",
       subtitle = "Check for homoscedasticity and patterns",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12))

# Plot 6: Combined diagnostic panel
p6_data <- data.frame(
  residuals = residuals(mod2),
  fitted = fitted(mod2),
  standardized = rstandard(mod2),
  index = 1:nrow(data)
)

p6a <- ggplot(p6_data, aes(sample = standardized)) +
  stat_qq(color = "steelblue", size = 2) +
  stat_qq_line(color = "red", linetype = "dashed", size = 1) +
  labs(title = "Normal Q-Q Plot", x = "Theoretical Quantiles", 
       y = "Standardized Residuals") +
  theme_minimal()

p6b <- ggplot(p6_data, aes(x = standardized)) +
  geom_histogram(aes(y = ..density..), bins = 15, fill = "steelblue", 
                 color = "white", alpha = 0.7) +
  geom_density(color = "darkblue", size = 1) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1),
                color = "red", linetype = "dashed", size = 1) +
  labs(title = "Residual Distribution", x = "Standardized Residuals", 
       y = "Density") +
  theme_minimal()

# Display all plots
print(p1)
print(p2)
print(p3)
print(p4)
print(p5)
grid.arrange(p6a, p6b, ncol = 2)

# Optional: Create a comprehensive 4-panel figure
combined_plot <- grid.arrange(p2, p3, p4, p5, ncol = 2, nrow = 2,
                               top = "Comprehensive Model Diagnostics for mod2")
print(combined_plot)
```



## Final thoughts

Multiple linear regression works in the same way than linear regression with a single variable. We rely on the same assumptions but we have the flexibility to add several continuous variables to explain our response variable.


## Session

```{r session, echo=TRUE}
sessionInfo()
```