---
title: "09 - Multiple linear regression"
author: "Francisco E. Fonturbel"
date: "09/August/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Multiple linear regression

Now that we know how to fit linear models, let's try something bolder. Why limiting us to a single predictor variable? In biological sciences, many phenomena depend on two or more variables than can be used to explain our response variable. Thus, the multivariate extension or linear regression is called **Multiple linear regression**. In this case, our linear model looks like this:

_y = a + b1x1 + b2x2 + b3x3 + ... + bnxn_

Being y the response variable, a and b's the coefficients, and x's the predictor variables in the model.

So, we will use the same `lm()` function but incorporating more than one predictor variable in the formula.

Let's use the `mtcars` dataset (included in R) to make an example of this

```{r data}
library(confintr)

data<-mtcars[,c("mpg","disp","hp","wt")]
attach(data)
head(data)
```

So, we have information on several car brands and models, obtained from their technical specifications. In this case

as

## Final thoughts

Linear regression and correlation are our entryway to understand linear relationships among linear variables. Linear regression allows us to assess the causal relationship between two continuous variables, whereas correlation examines association between a pair of variables without implying a causal relationship.


## Session

```{r session, echo=TRUE}
sessionInfo()
```