---
title: "11 - Generalized Linear Models and data distributions"
author: "Francisco E. Fonturbel"
date: "14/September/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Relaxing model assumptions... open many analysis possibilities

In the first part of this brief R course, we dealt with different tests that have similar assumptions (e.g., normality, homocedasticity, independence) but we know well that in real life many of those assumptions cannot be met using biological data. Also, we had some models (e.g., ANOVA) that use categorical predictors and others (e.g., linear regression) that use continuous predictors, but we were unable to mix both types of predictors in one model. Also, a major limitation of those _basic_ models is that we are stuck with a normal (= Gaussian) data distribution. But fortunately there are other options.

## Beyond the Gaussian distribution

Normal (= Gaussian) distribution is the entryway to parametric statistics, but there are many other options to consider. The Gaussian distribution works well in many cases, as it is a bell-shaped continuous distribution that can take negative and positive values. However, it is common to also have count data (e.g., counting insect individuals visiting a flower), which are integers (i.e., we cannot have half insect) and positive (no negative counts are possible). Or maybe we measure a response variable with a dichotomic response (success or failure, germinate or not germinate). In both cases, using a Gaussian distribution to analyse such kind of data is incorrect and it would lead to wrong conclusions.


## Generalized Linear Models in action

AAA

To illustrate this point, we will use data from [Fontúrbel et al. (2015)](https://www.mdpi.com/882552)

```{r data}
library(mgcv)
library(lme4)

data<-read.table("data/10_data_threat.txt", 
                     header=TRUE, 
                     sep="\t",
                     dec=",",
                     na.strings="NA")
attach(data)
head(data)
```


## Final thoughts

aaa


## Session

```{r session, echo=TRUE}
sessionInfo()
```