---
title: "12 - Generalized Estimation Equations and Special Distributions"
author: "Francisco E. Fonturbel"
date: "31/October/2025"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r packages, include=FALSE}
library(MASS)         # For glm.nb
library(lme4)         # For mixed models comparison
library(geepack)      # For GEE models
library(pscl)         # For zero-inflated models
library(ggplot2)      # For visualization
library(dplyr)        # For data manipulation
library(tidyr)        # For data reshaping
library(DHARMa)       # For residual diagnostics
library(AER)          # For dispersion tests
library(performance)  # For model comparison
```


# Relaxing Model Assumptions Even Further

Generalized Linear Models relax some assumptions (e.g., a normal distribution) and allow us to use count data or presence/absence data instead. In this opportunity, we will go beyond and relax other assumptions to properly analyze real life data.

## The Challenge with Real-World Data

Real ecological and biological data often violate standard GLM assumptions in two major ways:

1. **Excess zeros:** Many datasets have more zeros than expected by standard distributions
2. **Correlated observations:** Repeated measures or clustered data violate independence

### When Standard GLMs Fail

```{r conceptual_viz, fig.width=12, fig.height=4}
set.seed(123)
par(mfrow=c(1,3))

# 1. Normal Poisson
x_pois <- rpois(1000, lambda=3)
hist(x_pois, main="Standard Poisson Distribution", 
     xlab="Count", col="lightblue", breaks=20)

# 2. Zero-inflated Poisson
x_zip <- c(rep(0, 600), rpois(400, lambda=3))
hist(x_zip, main="Zero-Inflated Poisson\n(60% extra zeros)", 
     xlab="Count", col="salmon", breaks=20)

# 3. Overdispersed data
x_nb <- rnegbin(1000, mu=3, theta=1)
hist(x_nb, main="Overdispersed (Negative Binomial)", 
     xlab="Count", col="lightgreen", breaks=20)

par(mfrow=c(1,1))
```

**In this lesson we will:**
1. Compare regular GLMs with their zero-inflated alternatives
2. Use Generalized Estimation Equations (GEE) that relax the independence assumption
3. Learn how to diagnose and handle these data challenges

## Zero-Inflated Distributions

These are a special version of Poisson and negative binomial distributions in which we have an excess of zeros creating a deviation towards zero in the observed distribution. In many biological systems we face this problem, as capture/record events are rare. For example, when we set traps for capturing a given animal, in most cases the trapping success is lower than 20%. Thus, we will have a lot of zeros in our data, and therefore the conclusions from our GLM models may not be entirely reliable.

### Understanding Zero-Inflation

**Why do we get excess zeros?**
- **Structural zeros:** The event cannot occur (e.g., infertile individuals cannot reproduce)
- **Sampling zeros:** The event could occur but didn't during observation period
- **Detection zeros:** The event occurred but wasn't detected

**Zero-inflated models account for two processes:**
1. **Zero-inflation process:** Binary process determining if zero or non-zero
2. **Count process:** Standard count distribution (Poisson or Negative Binomial) for non-zeros

### Example Data: Biochemist Publications

In the past session we introduced different data distributions, but now we'll see zero-inflated models in action. To illustrate this we will use the _bioChemists_ dataset included in the `pscl` package, which contains publication counts for PhD biochemists.

```{r data}
data("bioChemists", package = "pscl")

# Examine data structure
str(bioChemists)
head(bioChemists, 10)

# Summary statistics
summary(bioChemists)
```

### Exploratory Data Analysis

```{r eda_biochem, fig.width=12, fig.height=8}
# Check for zero-inflation
n_zeros <- sum(bioChemists$art == 0)
n_total <- nrow(bioChemists)
prop_zeros <- n_zeros / n_total

print(paste("Number of zeros:", n_zeros, "out of", n_total))
print(paste("Proportion of zeros:", round(prop_zeros, 3)))

# What would Poisson predict?
mean_art <- mean(bioChemists$art)
expected_zeros_poisson <- dpois(0, lambda = mean_art) * n_total
print(paste("Expected zeros under Poisson:", round(expected_zeros_poisson, 1)))
print(paste("Observed zeros:", n_zeros))
print(paste("Excess zeros:", n_zeros - round(expected_zeros_poisson, 1)))

# Visualizations
par(mfrow=c(2,2))

# 1. Distribution of publications
hist(bioChemists$art, breaks=30, col="steelblue",
     main="Distribution of Article Counts",
     xlab="Number of Articles", ylab="Frequency")
abline(v=mean_art, col="red", lwd=2, lty=2)
text(mean_art + 1, max(table(bioChemists$art)) * 0.9, 
     paste("Mean =", round(mean_art, 2)), col="red")

# 2. Proportion of zeros
barplot(c(prop_zeros, 1-prop_zeros), 
        names.arg=c("Zero", "Non-zero"),
        col=c("coral", "lightgreen"),
        main="Proportion of Zeros vs Non-zeros",
        ylab="Proportion")

# 3. Publications by gender
boxplot(art ~ fem, data=bioChemists, 
        col=c("lightblue", "pink"),
        names=c("Men", "Women"),
        main="Publications by Gender",
        ylab="Number of Articles")

# 4. Publications by marital status
boxplot(art ~ mar, data=bioChemists,
        col=c("wheat", "lightgreen"),
        names=c("Single", "Married"),
        main="Publications by Marital Status",
        ylab="Number of Articles")

par(mfrow=c(1,1))

# Check variance vs mean
print(paste("Mean:", round(mean(bioChemists$art), 2)))
print(paste("Variance:", round(var(bioChemists$art), 2)))
print(paste("Variance/Mean ratio:", round(var(bioChemists$art)/mean(bioChemists$art), 2)))
print("Ratio > 1 suggests overdispersion")
```

### Model 1: Standard Poisson GLM

Now, let's make a regular GLM with this dataset:

```{r m1}
model1 <- glm(art ~ ., data = bioChemists, family = poisson)
summary(model1)

# Check for overdispersion
library(AER)
dispersion_test1 <- dispersiontest(model1)
print(dispersion_test1)

# Residual diagnostics
par(mfrow=c(2,2))
plot(model1)
par(mfrow=c(1,1))
```

Wow, we have a lot of significant terms in those results. However, let's check if this model is appropriate:

```{r m1_diagnostics}
# Check for overdispersion
print(paste("Residual deviance:", round(model1$deviance, 2)))
print(paste("Residual df:", model1$df.residual))
print(paste("Deviance/df ratio:", round(model1$deviance/model1$df.residual, 2)))
print("Ratio >> 1 indicates overdispersion")

# Formal test for overdispersion
if(dispersion_test1$p.value < 0.05) {
  print("✗ Significant overdispersion detected - standard Poisson may not be appropriate")
} else {
  print("✓ No significant overdispersion detected")
}

# Check observed vs predicted zeros
pred_prob_zero <- sum(dpois(0, fitted(model1)))
print(paste("Predicted number of zeros:", round(pred_prob_zero, 1)))
print(paste("Observed number of zeros:", sum(bioChemists$art == 0)))
```

### Model 2: Zero-Inflated Poisson (ZIP)

What about running this same dataset with a zero-inflated Poisson distribution instead:

```{r m1zi}
# The formula art ~ . | . means:
# Left side (art ~ .): Count model - predicts non-zero counts
# Right side (| .): Zero-inflation model - predicts excess zeros
model1zi <- zeroinfl(art ~ . | ., data = bioChemists, dist = "poisson")
summary(model1zi)

# Model interpretation
print("Count model coefficients: Effect on the count when non-zero")
print("Zero-inflation coefficients: Effect on probability of being a structural zero")
```

As we can see, some of the results change. The zero-inflated model separates two processes: what determines if someone publishes at all (zero-inflation part) versus how many articles they publish if they do publish (count part).

### Model 3: Negative Binomial GLM

Let's try the same models but using a negative binomial distribution instead of a Poisson distribution (to account for overdispersion):

```{r m2}
model2 <- glm.nb(art ~ ., data = bioChemists)
summary(model2)

print(paste("Theta (dispersion parameter):", round(model2$theta, 3)))
print("Larger theta = less overdispersion, closer to Poisson")

# Check residuals
par(mfrow=c(2,2))
plot(model2)
par(mfrow=c(1,1))
```

OK, some terms are significant. The negative binomial accounts for overdispersion through the theta parameter.

### Model 4: Zero-Inflated Negative Binomial (ZINB)

So, how would the zero-inflated version of this model look like?

```{r m2zi}
model2zi <- zeroinfl(art ~ . | ., data = bioChemists, dist = "negbin")
summary(model2zi)

print(paste("Theta:", round(model2zi$theta, 3)))
```

The results change and we have different significant factors in this case. Thus, using the correct distribution for our data is crucial to obtain reliable results.

### Comparing All Four Models

```{r model_comparison, fig.width=10, fig.height=6}
# AIC comparison
aic_values <- c(
  "Poisson" = AIC(model1),
  "ZIP" = AIC(model1zi),
  "Negative Binomial" = AIC(model2),
  "ZINB" = AIC(model2zi)
)

print("AIC values (lower is better):")
print(sort(aic_values))

#Comparsing ZI models:
print("Model Comparison: ZIP vs Poisson")
print(paste("Poisson AIC:", round(AIC(model1), 2)))
print(paste("ZIP AIC:", round(AIC(model1zi), 2)))
print(paste("AIC difference:", round(AIC(model1) - AIC(model1zi), 2)))
if(AIC(model1zi) < AIC(model1)) {
  print("→ Zero-inflated model is preferred (lower AIC)")
} else {
  print("→ Standard Poisson model is preferred (lower AIC)")
}

print("")
print("Model Comparison: ZINB vs Negative Binomial")
print(paste("Negative Binomial AIC:", round(AIC(model2), 2)))
print(paste("ZINB AIC:", round(AIC(model2zi), 2)))
print(paste("AIC difference:", round(AIC(model2) - AIC(model2zi), 2)))
if(AIC(model2zi) < AIC(model2)) {
  print("→ Zero-inflated model is preferred (lower AIC)")
} else {
  print("→ Standard Negative Binomial model is preferred (lower AIC)")
}

# Visualize model fit
par(mfrow=c(2,2))

# Observed vs predicted for each model
obs_freq <- table(factor(bioChemists$art, levels=0:max(bioChemists$art)))

# Model 1: Poisson
max_count <- max(bioChemists$art)
pred1 <- numeric(max_count + 1)
for(i in 0:max_count) {
  pred1[i+1] <- sum(dpois(i, fitted(model1)))
}
plot(0:max(bioChemists$art), obs_freq, type="h", lwd=3, col="black",
     main="Poisson GLM", xlab="Number of Articles", ylab="Frequency")
lines(0:max(bioChemists$art), pred1, col="red", lwd=2)
legend("topright", c("Observed", "Predicted"), col=c("black", "red"), lwd=2)

# Model 2: Negative Binomial
pred2 <- numeric(max_count + 1)
for(i in 0:max_count) {
  pred2[i+1] <- sum(dnbinom(i, mu=fitted(model2), size=model2$theta))
}
plot(0:max(bioChemists$art), obs_freq, type="h", lwd=3, col="black",
     main="Negative Binomial GLM", xlab="Number of Articles", ylab="Frequency")
lines(0:max(bioChemists$art), pred2, col="blue", lwd=2)
legend("topright", c("Observed", "Predicted"), col=c("black", "blue"), lwd=2)

# Model 3: ZIP
pred1zi <- predict(model1zi, type="prob")[,1:(max(bioChemists$art)+1)]
pred1zi_freq <- colSums(pred1zi)
plot(0:max(bioChemists$art), obs_freq, type="h", lwd=3, col="black",
     main="Zero-Inflated Poisson", xlab="Number of Articles", ylab="Frequency")
lines(0:max(bioChemists$art), pred1zi_freq, col="orange", lwd=2)
legend("topright", c("Observed", "Predicted"), col=c("black", "orange"), lwd=2)

# Model 4: ZINB
pred2zi <- predict(model2zi, type="prob")[,1:(max(bioChemists$art)+1)]
pred2zi_freq <- colSums(pred2zi)
plot(0:max(bioChemists$art), obs_freq, type="h", lwd=3, col="black",
     main="Zero-Inflated Negative Binomial", xlab="Number of Articles", ylab="Frequency")
lines(0:max(bioChemists$art), pred2zi_freq, col="purple", lwd=2)
legend("topright", c("Observed", "Predicted"), col=c("black", "purple"), lwd=2)

par(mfrow=c(1,1))
```

### Interpretation and Model Selection

```{r interpretation}
# Summary table
model_summary <- data.frame(
  Model = c("Poisson", "ZIP", "Neg Binomial", "ZINB"),
  AIC = round(aic_values, 1),
  Parameters = c(
    length(coef(model1)),
    length(coef(model1zi)),
    length(coef(model2)) + 1,  # +1 for theta
    length(coef(model2zi)) + 1
  )
)
model_summary$Delta_AIC <- model_summary$AIC - min(model_summary$AIC)
model_summary <- model_summary[order(model_summary$AIC),]

print("Model Selection Summary:")
print(model_summary)

print("")
print("Model Selection Guidelines:")
print("- Delta AIC < 2: Models are essentially equivalent")
print("- Delta AIC 2-7: Considerable support for best model")
print("- Delta AIC > 10: Best model is strongly preferred")
```

## Generalized Estimation Equations

There are many approaches to deal with non-normal data, but when our data has **correlated observations** (e.g., repeated measures, clustered data) we face additional challenges, as independence is one of the main assumptions of most statistical tests. Fortunately, there is an option to deal with correlated data: **Generalized Estimation Equations (GEE for short)**. 

### Understanding GEE

**When do we need GEE?**
- Repeated measurements on the same subjects over time
- Clustered/grouped observations (e.g., multiple observations per site)
- Spatial or temporal correlation in the data

**How GEE differs from GLM:**
- **GLM:** Assumes all observations are independent
- **GEE:** Accounts for correlation within clusters/subjects
- **Mixed Models:** Similar purpose but different approach (random effects vs correlation structure)

**Key concept:** GEE provides **population-averaged** estimates rather than subject-specific estimates (like mixed models).

### GEE Correlation Structures

GEE models require specifying a **working correlation structure**:

1. **Independence:** No correlation (like standard GLM)
2. **Exchangeable:** All observations within a cluster equally correlated
3. **AR(1) - Autoregressive:** Correlation decreases with time/distance
4. **Unstructured:** Estimate all correlations (requires more data)

```{r correlation_viz, fig.width=10, fig.height=3}
# Visualize correlation structures
par(mfrow=c(1,4))

# Independence
m_ind <- matrix(c(1,0,0,0, 0,1,0,0, 0,0,1,0, 0,0,0,1), 4, 4)
image(1:4, 1:4, m_ind, main="Independence", xlab="", ylab="", col=c("white", "steelblue"))

# Exchangeable
m_exch <- matrix(c(1,0.5,0.5,0.5, 0.5,1,0.5,0.5, 0.5,0.5,1,0.5, 0.5,0.5,0.5,1), 4, 4)
image(1:4, 1:4, m_exch, main="Exchangeable", xlab="", ylab="", col=heat.colors(20))

# AR(1)
m_ar1 <- matrix(c(1,0.7,0.49,0.34, 0.7,1,0.7,0.49, 0.49,0.7,1,0.7, 0.34,0.49,0.7,1), 4, 4)
image(1:4, 1:4, m_ar1, main="AR(1)", xlab="", ylab="", col=heat.colors(20))

# Unstructured (example)
m_unstr <- matrix(c(1,0.6,0.4,0.3, 0.6,1,0.7,0.5, 0.4,0.7,1,0.6, 0.3,0.5,0.6,1), 4, 4)
image(1:4, 1:4, m_unstr, main="Unstructured", xlab="", ylab="", col=heat.colors(20))

par(mfrow=c(1,1))
```

### Example Data: Pig Weight Gain

To illustrate this point we will use the _dietox_ example data file from the `geepack` package. This dataset tracks pig weight over time under different copper supplementation treatments.

```{r gee_data}
data(dietox)

# Convert Cu to factor
dietox$Cu <- as.factor(dietox$Cu)

# Examine structure
str(dietox)
head(dietox, 15)

# Summary
print("Number of pigs:")
length(unique(dietox$Pig))

print("Observations per pig:")
table(table(dietox$Pig))

print("Time points:")
sort(unique(dietox$Time))
```

### Exploratory Analysis

```{r gee_eda, fig.width=12, fig.height=8}
# Visualize individual trajectories
ggplot(dietox, aes(x=Time, y=Weight, group=Pig, color=Cu)) +
  geom_line(alpha=0.3) +
  geom_smooth(aes(group=Cu), method="loess", se=TRUE, size=1.5) +
  facet_wrap(~Cu, labeller=label_both) +
  labs(title="Pig Weight Trajectories by Copper Treatment",
       x="Time (weeks)", y="Weight (kg)") +
  theme_minimal() +
  theme(legend.position="bottom")

# Summary statistics by treatment and time
summary_stats <- dietox %>%
  group_by(Cu, Time) %>%
  summarise(
    n = n(),
    mean_weight = mean(Weight),
    sd_weight = sd(Weight),
    se_weight = sd(Weight)/sqrt(n()),
    .groups = 'drop'
  )

print("Summary by copper treatment and time:")
print(head(summary_stats, 12))

# Check correlation within pigs
sample_pigs <- sample(unique(dietox$Pig), 4)
par(mfrow=c(2,2))
for(pig in sample_pigs) {
  pig_data <- dietox[dietox$Pig == pig,]
  plot(pig_data$Time, pig_data$Weight, type="b", 
       main=paste("Pig", pig), xlab="Time", ylab="Weight",
       pch=16, col="steelblue")
}
par(mfrow=c(1,1))
```

### Fitting GEE Models

GEE models allow us to indicate a correlation structure and obtain reliable estimates accounting for the repeated measures on the same pigs.

```{r geemod}
# Define model formula
# Weight as polynomial function of time, with copper treatment interaction
mf <- formula(Weight ~ Cu * (Time + I(Time^2) + I(Time^3)))

# Fit GEE with AR(1) correlation structure
gee1 <- geeglm(mf, data=dietox, id=Pig, family=poisson("identity"), corstr="ar1")
summary(gee1)

print("Estimated correlation parameter (alpha):")
print(gee1$geese$alpha)
print("This represents the correlation between adjacent time points")
```

As you can see, GEE models operate in a similar way to GLM models and produce similar outputs as well.

### Comparing Different Correlation Structures

```{r gee_corstr, fig.width=10, fig.height=6}
# Fit models with different correlation structures
gee_ind <- geeglm(mf, data=dietox, id=Pig, family=poisson("identity"), corstr="independence")
gee_exch <- geeglm(mf, data=dietox, id=Pig, family=poisson("identity"), corstr="exchangeable")
gee_ar1 <- geeglm(mf, data=dietox, id=Pig, family=poisson("identity"), corstr="ar1")

# Compare using QIC (quasi-likelihood information criterion)
qic_values <- c(
  "Independence" = QIC(gee_ind)[1],
  "Exchangeable" = QIC(gee_exch)[1],
  "AR(1)" = QIC(gee_ar1)[1]
)

print("QIC values (lower is better):")
print(qic_values)

barplot(qic_values, col="steelblue", 
        main="Model Comparison by QIC",
        ylab="QIC Value")

# Extract correlation parameters
print("Estimated correlation parameters:")
print(paste("Exchangeable alpha:", round(gee_exch$geese$alpha, 3)))
print(paste("AR(1) alpha:", round(gee_ar1$geese$alpha, 3)))
```

### GEE vs GLM vs Mixed Models

```{r gee_comparison}
# Standard GLM (ignoring correlation)
glm_model <- glm(mf, data=dietox, family=poisson("identity"))

# Mixed model with random intercept (for comparison)
lmer_model <- lmer(Weight ~ Cu * (Time + I(Time^2) + I(Time^3)) + (1|Pig), 
                   data=dietox)

# Compare coefficient estimates
coef_comparison <- data.frame(
  Parameter = names(coef(gee_ar1)),
  GEE_AR1 = coef(gee_ar1),
  GLM = coef(glm_model),
  Mixed_Model = fixef(lmer_model)
)
print("Coefficient Comparison:")
coef_comparison[,2:4] <- round(coef_comparison[,2:4], 4)
print(coef_comparison)

# Compare standard errors
se_gee <- summary(gee_ar1)$coefficients[,"Std.err"]
se_glm <- summary(glm_model)$coefficients[,"Std. Error"]
se_lmer <- sqrt(diag(vcov(lmer_model)))

se_comparison <- data.frame(
  Parameter = names(coef(gee_ar1)),
  SE_GEE = se_gee,
  SE_GLM = se_glm,
  SE_Mixed = se_lmer
)

print("Standard Error Comparison:")
se_comparison[,2:4] <- round(se_comparison[,2:4], 4)
print(se_comparison)
print("")
print("Note: GEE and Mixed Models typically have larger SE than GLM")
print("because they properly account for correlation")
```

### Visualizing GEE Results

```{r gee_viz, fig.width=12, fig.height=6}
# Create prediction data
pred_data <- expand.grid(
  Cu = levels(dietox$Cu),
  Time = seq(min(dietox$Time), max(dietox$Time), length.out=50)
)

# Get predictions from GEE model
pred_data$Weight <- predict(gee_ar1, newdata=pred_data, type="response")

# Plot predictions with observed data
ggplot() +
  geom_point(data=dietox, aes(x=Time, y=Weight, color=Cu), alpha=0.2) +
  geom_line(data=pred_data, aes(x=Time, y=Weight, color=Cu), size=1.5) +
  facet_wrap(~Cu, labeller=label_both) +
  labs(title="GEE Model Predictions vs Observed Data",
       subtitle="Polynomial growth curves with AR(1) correlation structure",
       x="Time (weeks)", y="Weight (kg)") +
  theme_minimal() +
  theme(legend.position="bottom")
```

### Model Diagnostics for GEE

```{r gee_diagnostics}
# Residuals
residuals_gee <- residuals(gee_ar1, type="pearson")

# Diagnostic plots
par(mfrow=c(2,2))

# Residuals vs fitted
plot(fitted(gee_ar1), residuals_gee,
     xlab="Fitted Values", ylab="Pearson Residuals",
     main="Residuals vs Fitted", pch=16, col=rgb(0,0,1,0.3))
abline(h=0, col="red", lty=2)

# Q-Q plot
qqnorm(residuals_gee, main="Normal Q-Q Plot", pch=16, col=rgb(0,0,1,0.3))
qqline(residuals_gee, col="red")

# Residuals by time
plot(dietox$Time, residuals_gee,
     xlab="Time", ylab="Pearson Residuals",
     main="Residuals vs Time", pch=16, col=rgb(0,0,1,0.3))
abline(h=0, col="red", lty=2)

# Residuals by treatment
boxplot(residuals_gee ~ dietox$Cu,
        main="Residuals by Treatment",
        xlab="Copper Treatment", ylab="Pearson Residuals",
        col="lightblue")
abline(h=0, col="red", lty=2)

par(mfrow=c(1,1))
```

## When to Use Each Approach

### Decision Tree for Count Data

```{r decision_tree, echo=FALSE}
cat("
DECISION TREE FOR COUNT DATA ANALYSIS
======================================

1. Do you have COUNT data? (0, 1, 2, 3, ...)
   YES → Continue
   NO → Use appropriate GLM family (binomial, gamma, etc.)

2. Are observations INDEPENDENT?
   NO (repeated measures, clustered) → Use GEE or Mixed Models
   YES → Continue

3. Is there OVERDISPERSION? (Variance >> Mean)
   NO → Try Poisson GLM
   YES → Continue

4. Are there EXCESS ZEROS?
   NO → Use Negative Binomial GLM
   YES → Continue

5. Use ZERO-INFLATED models:
   - If mild overdispersion → Zero-Inflated Poisson (ZIP)
   - If strong overdispersion → Zero-Inflated Negative Binomial (ZINB)

For correlated data with excess zeros:
→ Consider zero-inflated mixed models (not covered here)
")
```

### Summary Table

```{r summary_table, echo=FALSE}
comparison_table <- data.frame(
  Method = c("Poisson GLM", "Negative Binomial", "ZIP", "ZINB", "GEE", "Mixed Models"),
  `Use When` = c(
    "Count data, no overdispersion, independent obs",
    "Count data, overdispersion, independent obs",
    "Excess zeros, mild overdispersion, independent",
    "Excess zeros, strong overdispersion, independent",
    "Correlated observations, population-level inference",
    "Correlated observations, subject-level inference"
  ),
  Assumption = c(
    "Independence, variance=mean",
    "Independence, variance>mean",
    "Independence, two processes",
    "Independence, two processes + overdispersion",
    "Specified correlation structure",
    "Random effects structure"
  ),
  stringsAsFactors = FALSE
)

print("METHOD SELECTION GUIDE:")
print(comparison_table)
```

## Final thoughts

Relaxing other assumptions allows us to properly analyze other kinds of data. On the one hand, **zero-inflated models** allow us to fit Poisson and negative binomial GLMs when we have an excess of zeros in our dataset, correcting the estimates and P-values by modeling two separate processes: one determining whether a zero occurs, and another determining the count when non-zero. On the other hand, **GEE** works in a similar way to GLMs but is able to operate with correlated observations (such as repeated measures or clustered data), which represent a major assumption violation in standard linear models.

### Key Takeaways:

**Zero-Inflated Models:**
- Essential when you have more zeros than expected by standard distributions
- Model two processes: zero-inflation and count generation
- Use Vuong test to compare zero-inflated vs standard models
- Choose between ZIP and ZINB based on overdispersion

**Generalized Estimation Equations:**
- Appropriate for correlated/clustered data
- Provide population-averaged (marginal) estimates
- Require specification of correlation structure
- More robust to misspecification than mixed models
- Use QIC for model selection

**Best Practices:**
1. Always visualize your data first
2. Check for overdispersion using formal tests
3. Compare multiple models using AIC/QIC
4. Validate model fit with diagnostic plots
5. Choose correlation structure based on data structure
6. Report model selection process in your results

**When in doubt:**
- Start with exploratory data analysis
- Fit multiple candidate models
- Use information criteria for comparison
- Check model assumptions
- Validate predictions against observed data

## Session

```{r session, echo=TRUE}
sessionInfo()
```