---
title: "12 - Generalized Estimation Equations and special distributions"
author: "Francisco E. Fonturbel"
date: "26/September/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# Relaxing model assumptions even further

Generalized Linear Models relax some assumptions (e.g., a normal distribution) and allow us to use count data or presence/absence data instead. In this opportunity, we will go beyond and relax other assumptions to properly analyze real life data.

In this opportunity, we will (1) compare regular GLMs with their zero-inflated alter-egos, and (2) use Generalized Estimation Equations (GEE) that are a variation of GLMs relaxing the assumption of variance homogeneity.


## Zero-inflated distributions

These are an special version of Poisson and negative binomial distributions in which we have an excess of zeroes creating a deviation towards zero in the observed distribution. In many biological systems we face this problem, as capture / record events are rare. For example, when we set traps for capturing a given animal, in most cases the trapping success is lower than 20%. Thus, we will have a lot of zeroes in our data, and therefore the conclusions from our GLM models may not be entirely reliable.

In the past session we introduced different data distributions, but now we'll see zero-inflated models in action. To illustrate this we will use the _bioChemists_ dataset included in the `pscl` package.

```{r data}
library(MASS)
library(lme4)
library(geepack)
library(pscl)

data("bioChemists", package = "pscl")
```

Now, let's make a regular GLM with this dataset:

```{r m1}
model1 <- glm(art ~ ., data = bioChemists, family = poisson)
summary(model1)
```

Wow, we have a lot of significant terms in those results.

What about running this same dataset with a zero-inflated Poisson distribution instead:

```{r m1zi}
model1zi <- zeroinfl(art ~ . | ., data = bioChemists, dist = "poisson")
summary(model1zi)
```

As we can see, some of the results change. Let's try the same models but using a negative binomial distribution instead of a Poisson distribution:

```{r m2}
model2 <- glm.nb(art ~ ., data = bioChemists)
summary(model2)
```

OK, some terms are significant. So, how the zero-inflated version of this model would look like?

```{r m2zi}
model2zi <- zeroinfl(art ~ . | ., data = bioChemists, dist = "negbin")
summary(model2zi)
```

The results change and we have only one significant factor in this case. Thus, using the correct distribution to our data is crucial to obtain reliable results.


## Generalized Estimation Equations

There are many approaches to deal with non-normal data, but when our data has heterogeneous variances we are quite swamped with our data analysis, as homocedasticity is one of the main assumptions of most statistical tests. Fortunately, there is an option to deal with heterocedasticity: Generalized Estimation Equations (GEE for short). GEE models allow us to indicate a correlation structure and obtain reliable estimates.

To illustrate this point we will use the _dietox_ example data file from the `geepack` package.

```{r geemod}
data(dietox)
dietox$Cu     <- as.factor(dietox$Cu)
mf <- formula(Weight ~ Cu * (Time + I(Time^2) + I(Time^3)))
gee1 <- geeglm(mf, data=dietox, id=Pig, family=poisson("identity"), corstr="ar1")
summary(gee1)
```

As you can see, GEE models operate in a similar way than GLM models and produce similar outputs as well.


## Final thoughts

Relaxing other assumptions allow us to properly analyze other kind of data. On the one hand, zero-inflated models allow to fit Poisson and negative binomial GLMs when we have an excess of zeroes in out dataset, correcting the estimates and P-values. On the other hand, GEE work in a similar way than GLMs but are able to operate with heterogeneous variances, which represent a major assumption violation in most linear models.


## Session

```{r session, echo=TRUE}
sessionInfo()
```