---
title: "16 - ANOSIM and SIMPER"
author: "Francisco E. Fonturbel"
date: "23/November/2022"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## How to analyze species composition?

Dealing with biodiversity data is somewhat different than dealing with the usual stuff that we are used to. Some authors use standard approaches (e.g., ANOVA) to compare species composition among sites, but this is not correct as in this case the identity of the species matters, and traditional methods cannot consider them.

```{r data}
library(vegan)

data<-read.csv("data/16_bryophytes.csv", header = T, sep = ";")
attach(data)
head(data)
```

We will use the data of [Riffo-Donoso et al. 2021](https://www.sciencedirect.com/science/article/pii/S0378112721005569) to illustrate this. This dataset contains bryophyte presence-absence data for three sampling years, comparing four habitat conditions and four substrates.

## Analysis of Similarities (ANOSIM)

When we have species presence/absence or relative abundance data with two or more groups and at least three replicates per group, we can use the Analysis of Similarities (ANOSIM) to test for spècies composition differences. ANOSIM is a non-parametric approach that uses a similarity metric (e.g., Jaccard or Bray-Curtis)


## Final thoughts

Linear regression and correlation are our entryway to understand linear relationships among linear variables. Linear regression allows us to assess the causal relationship between two continuous variables, whereas correlation examines association between a pair of variables without implying a causal relationship.


## Session

```{r session, echo=TRUE}
sessionInfo()
```