---
title: "16 - ANOSIM, SIMPER, and mvabund for Community Analysis"
author: "Francisco E. Fonturbel"
date: "05/October/2026"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## How to analyze species composition?

Dealing with biodiversity data is somewhat different than dealing with the usual stuff that we are used to. Some authors use standard approaches (e.g., ANOVA) to compare species composition among sites, but this is not correct as in this case the identity of the species matters, and traditional methods cannot consider them.

### Understanding the Challenge

Traditional univariate tests (like ANOVA) have several limitations when applied to community data:

1. **Ignore species identities:** Treating each species independently loses the multivariate structure
2. **Multiple testing problem:** Testing each species separately inflates Type I error
3. **Ignore correlations:** Species often co-vary due to similar habitat requirements
4. **Compositional nature:** Community data has special properties (many zeros, over-dispersion)

### Loading and Exploring the Data

```{r data}
#Loading packages
library(vegan)      # For ANOSIM and SIMPER
library(mvabund)    # For multivariate abundance analysis
library(ggplot2)    # For advanced plotting
library(dplyr)      # For data manipulation
library(tidyr)      # For data reshaping
library(corrplot)   # For correlation plots
library(nestedANOSIM) # For two-way ANOSIM and SIMPER (github.com/fonturbel/nestedANOSIM)

data<-read.table("data/16_bird_data.txt", 
                     header=TRUE, 
                     sep="\t", dec=".", 
                     na.strings="NA")

# Check data structure
str(data)
head(data)

# Summary of experimental design
print("Sample sizes by habitat:")
table(data$Habitat)

print("Sample sizes by sampling method:")
table(data$Sample)

print("Cross-tabulation:")
table(data$Sample, data$Habitat)
```

We will use the data of [Fontúrbel et al. 2020](https://www.sciencedirect.com/science/article/pii/S1470160X20308013) to illustrate this. This dataset contains bird abundance data for camera trap and point count data, comparing four habitat conditions.

### Data Exploration

```{r exploration}
# Extract species data
birds<-data[,3:22]

# Basic summaries
print("Total abundance per species:")
colSums(birds)

print("Total abundance per site:")
rowSums(birds)

# Species richness per site
richness <- rowSums(birds > 0)
print("Species richness summary:")
summary(richness)

# Visualize total abundances
par(mar=c(8,4,2,1))
barplot(colSums(birds), las=2, col="steelblue",
        main="Total Abundance by Species",
        ylab="Total Count")
```

### Visualizing Community Composition

```{r composition_viz, fig.width=12, fig.height=8}
# Prepare data for stacked barplot
birds_with_groups <- cbind(data[,1:2], birds)

# Get the species column names (all columns except Sample and Habitat)
species_cols <- names(birds)

# Calculate proportions by habitat
habitat_composition <- birds_with_groups %>%
  group_by(Habitat) %>%
  summarise(across(all_of(species_cols), sum)) %>%
  pivot_longer(cols = -Habitat, names_to = "Species", values_to = "Abundance")

# Stacked barplot by habitat
ggplot(habitat_composition, aes(x=Habitat, y=Abundance, fill=Species)) +
  geom_bar(stat="identity", position="fill") +
  labs(title="Bird Community Composition by Habitat",
       y="Proportion of Total Abundance") +
  theme_minimal() +
  theme(legend.position="right")

# Calculate proportions by sampling method
sample_composition <- birds_with_groups %>%
  group_by(Sample) %>%
  summarise(across(all_of(species_cols), sum)) %>%
  pivot_longer(cols = -Sample, names_to = "Species", values_to = "Abundance")

# Stacked barplot by sampling method
ggplot(sample_composition, aes(x=Sample, y=Abundance, fill=Species)) +
  geom_bar(stat="identity", position="fill") +
  labs(title="Bird Community Composition by Sampling Method",
       y="Proportion of Total Abundance") +
  theme_minimal() +
  theme(legend.position="right")
```

## Analysis of Similarities (ANOSIM)

When we have species presence/absence or relative abundance data with two or more groups and at least three replicates per group, we can use the Analysis of Similarities (ANOSIM) to test for species composition differences. ANOSIM is a non-parametric approach that uses a similarity metric (e.g., Jaccard or Bray-Curtis) to compare species similarity within and between groups (following the same logic than ANOVA). 

### Understanding ANOSIM

**How ANOSIM works:**
1. Calculate a dissimilarity matrix between all pairs of samples
2. Rank all dissimilarities
3. Compare average ranks within groups vs. between groups
4. Test significance using permutations

**The R statistic:**
- R ranges from -1 to +1
- R ≈ 0: No difference between groups
- R = 1: Complete separation (all within-group dissimilarities < all between-group)
- R < 0: Within-group dissimilarities > between-group (rare, suggests misclassification)

**Interpretation guidelines:**
- R > 0.75: Well separated
- R > 0.5: Separated but overlapping
- R > 0.25: Separated but strongly overlapping
- R < 0.25: Barely separable

The resulting statistic (R) is not a proportional value as it is calculated upon rankings (and it should not be interpreted as a linear scale of difference). The significance of this test is estimated using permutations (P value is the number of times that the randomized R value is equal or greater than the real R value, divided by the number of permutations). ANOSIM is implemented in the `vegan` package.

### ANOSIM: Testing Habitat Effects

```{r anosim1}
#Creating a distance matrix
birdist<-vegdist(birds, method="bray")

# Check the distribution of dissimilarities
hist(birdist, main="Distribution of Bray-Curtis Dissimilarities",
     xlab="Dissimilarity", col="lightblue", breaks=20)

#ANOSIM per habitat
# birdist is already a Bray-Curtis matrix, so anosim() needs no 'distance' argument
forestbirds<-anosim(birdist, data$Habitat, permutations = 999)
summary(forestbirds)
plot(forestbirds, xlab="Condition", ylab="Rank distance")

# Additional information
print(paste("ANOSIM R statistic:", round(forestbirds$statistic, 3)))
print(paste("Significance:", forestbirds$signif))
print("Interpretation: R < 0.25 suggests communities are barely separable by habitat")
```

So, there is no significant difference among habitat conditions (R = `r round(forestbirds$statistic, 3)`, p = `r forestbirds$signif`). This suggests that bird communities are similar across the different habitat types sampled.

### ANOSIM: Testing Sampling Method Effects

What about sampling approaches?

```{r anosim2}
sampbirds<-anosim(birdist, data$Sample, permutations = 999)
summary(sampbirds)
plot(sampbirds, xlab="Sampling method", ylab="Rank distance")

print(paste("ANOSIM R statistic:", round(sampbirds$statistic, 3)))
print(paste("Significance:", sampbirds$signif))
print("Interpretation: R > 0.25 suggests some separation between sampling methods")
```

In this case, we found a significant difference (R = `r round(sampbirds$statistic, 3)`, p = `r sampbirds$signif`), in which point counts are different from camera traps. This makes biological sense as different sampling methods may detect different species or relative abundances.

### Two-way ANOSIM: habitat within each sampling method

The two tests above look at one factor at a time, but our design has two: **habitat** (4 levels) and **sampling method** (3 levels), with 3 sites per habitat. Both factors are **crossed**: every habitat was sampled with every method. Since sampling method has a strong effect, it could be hiding a habitat effect. A two-way ANOSIM answers this by testing sampling method first and then habitat *within* each sampling method separately.

`vegan` only implements one-way ANOSIM, so we will use the `nested_anosim()` function from the `nestedANOSIM` package (install it with `remotes::install_github("fonturbel/nestedANOSIM")`). It runs three tests on the same Bray-Curtis matrix:

1. An overall test on the combined method x habitat groups.
2. A test of the main factor (sampling method).
3. A test of the second factor (habitat) within each level of the main factor, with a correction for multiple tests (`p_adjust`).

There is one more detail. The **same 12 sites** were sampled with point counts and, in two consecutive years, with camera traps (`Camera1` = first year, `Camera2` = second year), so the three samples from a site are not independent (they are *paired*). To account for this, we restrict the permutations of the sampling-method test to happen *within* each site using the `strata` argument. Within a single method each site appears only once, so the habitat tests use ordinary permutations.

```{r anosim_twoway}
# Site ID: the same 12 sites (rows 1-12) were sampled with each method
site <- paste0("S", rep(1:12, 3))
table(site, data$Sample)[1:3, ]   # each site appears once per method

birds2 <- nested_anosim(comm   = birds,
                        main   = data$Sample,    # main factor
                        nested = data$Habitat,   # tested within each method
                        strata = site,           # paired design: permute within sites
                        method = "bray",
                        permutations = 9999,
                        p_adjust = "holm",       # we run 3 habitat tests
                        seed = 123)
```

The effect of sampling method holds after accounting for the paired design (R = `r round(birds2$main$statistic, 3)`, p = `r format(birds2$main$signif, scientific = FALSE)`). However, habitat has no effect even when we compare habitats within the same sampling method:

```{r anosim_twoway_table}
birds2$nested
```

Notice the **negative R** within `Camera2` (R = `r round(birds2$nested$R[birds2$nested$main_level == "Camera2"], 3)`): dissimilarities among sites of the same habitat are *larger* than those between habitats. Also keep in mind that we only have 3 sites per habitat and method, so these tests have little statistical power: a non-significant result does not prove that habitats are identical.

The `plot_anosim_box()` function draws the ANOSIM boxplots with `ggplot2`. It accepts a two-way result (choosing the test with `which`) or a regular `anosim` object:

```{r anosim_twoway_plots, fig.height=4.5, fig.width=6}
plot_anosim_box(birds2, which = "main", xlab = "Sampling method")
plot_anosim_box(birds2, which = "Point", xlab = "Habitat")
```

Which factor goes first depends on the question. Here we asked whether habitat matters once we control for the sampling method. If we wanted to know whether the sampling methods differ within each habitat, we would use `main = data$Habitat` and `nested = data$Sample` instead.

```{r anosim_ordination, eval=FALSE, fig.height=6, fig.width=12, include=FALSE}
### Visualizing ANOSIM Results with Ordination [temporarily disabled]

# NMDS ordination to visualize community differences
set.seed(123)
nmds <- metaMDS(birds, distance="bray", k=2, trymax=100)

# Extract NMDS scores
nmds_scores <- as.data.frame(scores(nmds, display="sites"))
nmds_scores$Habitat <- data$Habitat
nmds_scores$Sample <- data$Sample

# Plot by habitat
par(mfrow=c(1,2))
plot(nmds, type="n", main="NMDS by Habitat")
points(nmds, display="sites", col=as.numeric(factor(data$Habitat)), pch=16, cex=1.5)
legend("topright", legend=levels(factor(data$Habitat)), 
       col=1:nlevels(factor(data$Habitat)), pch=16)
ordihull(nmds, groups=data$Habitat, draw="polygon", alpha=50, 
         col=1:nlevels(factor(data$Habitat)))

# Plot by sampling method
plot(nmds, type="n", main="NMDS by Sampling Method")
points(nmds, display="sites", col=as.numeric(factor(data$Sample)), pch=16, cex=1.5)
legend("topright", legend=levels(factor(data$Sample)), 
       col=1:nlevels(factor(data$Sample)), pch=16)
ordihull(nmds, groups=data$Sample, draw="polygon", alpha=50,
         col=1:nlevels(factor(data$Sample)))

par(mfrow=c(1,1))

print(paste("NMDS stress:", round(nmds$stress, 3)))
print("Stress interpretation: < 0.1 excellent, < 0.2 good, > 0.3 poor")
```

## Which species explain the differences?

In the second case, we found a significant difference in the ANOSIM test. But, which species account for such differences? In this case, we can use the Similarity Percentages (SIMPER) test, also implemented in the `vegan` package.

### Understanding SIMPER

**SIMPER (Similarity Percentages) analysis:**
- Identifies species contributing most to observed differences between groups
- Calculates average contribution of each species to dissimilarity
- Provides cumulative contribution percentages
- Helps focus on key species driving community patterns

```{r simper}
simbirds<-simper(birds, group = data$Sample, permutations = 999)
summary(simbirds, digits = 3)


# The comparisons are named after the groups, in order of appearance in the data
names(simbirds)
```

The output of `summary()` is a list with one table per pair of groups, which is not very convenient to work with. The `nested_simper()` function from `nestedANOSIM` returns the same results as a single tidy data frame. It uses the same design as the two-way ANOSIM: the main factor (sampling method, permuting within sites) and the second factor (habitat) within each sampling method. By default it keeps the species that together explain 70% of the dissimilarity (`cutoff = 0.7`):

```{r simper_twoway}
simbirds2 <- nested_simper(comm   = birds,
                           main   = data$Sample,
                           nested = data$Habitat,
                           strata = site,          # paired design, as in the ANOSIM
                           permutations = 999,
                           cutoff = 0.7,
                           seed = 123)

# Tidy table for the sampling methods: one row per species and comparison
simbirds2$main[, c("comparison", "species", "pct", "cum_pct", "mean_a", "mean_b", "p")]
```

`pct` is the percentage of the average dissimilarity explained by each species, `cum_pct` is the cumulative percentage, and `mean_a` and `mean_b` are the mean abundances in the first and second group of each comparison. Being a data frame, it is easy to plot:

```{r simper_twoway_plot, fig.height=4.5, fig.width=7}
top <- subset(simbirds2$main, comparison == "Point vs Camera1")
barplot(top$pct, names.arg = top$species, las = 2, col = "coral",
        main = "Species contributing to dissimilarity\n(Point vs Camera1)",
        ylab = "Contribution (%)", cex.names = 0.7)
```

The hummingbird (`COLIBRI`) alone explains about `r round(simbirds2$main$pct[simbirds2$main$comparison == "Point vs Camera1" & simbirds2$main$species == "COLIBRI"])`% of the difference between point counts and camera traps, as both methods detect it very differently.

SIMPER p-values must be read with care when there are more than two groups. To compute them, `vegan` permutes the labels of *all* groups, so the p-value of one pair is also affected by the other groups. For instance, several species look "significant" between the two camera years:

```{r simper_cameras}
cams <- subset(simbirds2$main, comparison == "Camera1 vs Camera2")
cams[, c("species", "pct", "mean_a", "mean_b", "p")]
```

However, a pairwise ANOSIM shows that the two camera years do not differ as a whole:

```{r anosim_cameras}
two_cams <- data$Sample != "Point"
set.seed(123)
cameras <- anosim(vegdist(birds[two_cams, ], method = "bray"),
                  data$Sample[two_cams], permutations = 9999,
                  strata = site[two_cams])
cameras
```

The bird communities recorded by the cameras did not differ between years (R = `r round(cameras$statistic, 3)`, p = `r round(cameras$signif, 3)`), so camera traps produced consistent results from one year to the next. The species-level p-values from SIMPER are therefore better used as a guide than as formal tests.

Keep in mind that SIMPER contributions are descriptive: abundant and highly variable species tend to dominate them (Warton et al. 2012, *Methods in Ecology and Evolution* 3: 89-101). This is one of the reasons to use model-based approaches such as `mvabund`, which we will see next.

## Multivariate Abundance Analysis using mvabund

While ANOSIM and SIMPER are useful, they have some limitations:
- **ANOSIM:** Assumes exchangeability, sensitive to dispersion differences, uses rank-based approach
- **Limited inference:** Doesn't provide species-level p-values or account for mean-variance relationships

The `mvabund` package provides a more powerful alternative based on **multivariate generalized linear models (GLMs)** that:
- **Tests community-level hypotheses** while accounting for species correlations
- **Provides species-level inference** with proper p-value adjustment
- **Handles mean-variance relationships** common in count data
- **More powerful** for detecting subtle community differences
- **Model-based approach** allowing for more complex experimental designs

### Understanding mvabund

**Key advantages of mvabund:**

1. **Proper statistical framework:** Uses likelihood-based models (GLMs) rather than distance-based approaches
2. **Accounts for mean-variance relationship:** Count data typically shows increasing variance with mean
3. **Species correlations:** Resampling approach accounts for correlations among species
4. **Hypothesis testing:** Provides proper multivariate test statistics
5. **Species-level inference:** Identifies which species contribute to differences with adjusted p-values
6. **Flexible:** Can include covariates and more complex designs

### Preparing Data for mvabund

```{r mvabund_setup}
# Create mvabund object
birds_mv <- mvabund(birds)

## Check the mean-variance relationship
# Alternative approach if meanvar.plot gives issues:
species_means <- colMeans(birds)
species_vars <- apply(birds, 2, var)

plot(species_means, species_vars, 
     pch=16, col="steelblue",
     xlab="Mean Abundance", 
     ylab="Variance",
     main="Mean-Variance Relationship")
abline(a=0, b=1, col="red", lty=2)  # 1:1 line
text(max(species_means)*0.7, max(species_vars)*0.9, 
     "If points are above the line,\nvariance > mean (overdispersion)", 
     cex=0.8)

# The plot shows if variance increases with mean (common in count data)
# If variance increases with mean, negative binomial family is appropriate
```

### Choosing the Right Distribution Family

```{r model_selection}
# Compare different families for our data
# This helps us choose the appropriate error distribution

# Negative binomial is often best for count data with overdispersion
# Let's check by fitting models and comparing

# Fit models with different families
mod_poisson <- manyglm(birds_mv ~ data$Sample, family="poisson")
mod_negbin <- manyglm(birds_mv ~ data$Sample, family="negative.binomial")

# Plot residuals to check assumptions
par(mfrow=c(2,2))
plot(mod_poisson, which=1:2, main="Poisson")
plot(mod_negbin, which=1:2, main="Negative Binomial")
par(mfrow=c(1,1))

print("Based on residual plots, negative binomial typically fits better for count data")
print("We'll use negative binomial family for subsequent analyses")
```

### mvabund Analysis: Testing Sampling Method Effects

```{r mvabund_sample}
# Fit the model
mv_sample <- manyglm(birds_mv ~ data$Sample, family="negative.binomial")

# Summary of the model
summary(mv_sample)

# Hypothesis test using likelihood ratio test with resampling
# This tests if bird community composition differs by sampling method
mv_sample_test <- anova(mv_sample, 
                        p.uni="adjusted",  # Get adjusted p-values for each species
                        nBoot=999,         # Number of bootstrap iterations
                        test="LR")         # Likelihood ratio test

print("Multivariate test results:")
print(mv_sample_test)

# Interpret results
pval_sample <- mv_sample_test$table["data$Sample", "Pr(>Dev)"]
if(!is.na(pval_sample) && pval_sample < 0.05) {
  print("✓ Significant difference in community composition by sampling method")
} else if(!is.na(pval_sample)) {
  print("✗ No significant difference in community composition by sampling method")
} else {
  print("Note: P-value not available - check test results above")
}
```

### Species-Level Inference with mvabund

```{r species_level}
# Get species-level p-values
species_pvals <- mv_sample_test$uni.p[2,]  # p-values for Sample effect
species_test <- mv_sample_test$uni.test[2,]  # Test statistics

# Create results table
species_results <- data.frame(
  Species = names(species_pvals),
  Test_Stat = species_test,
  P_value = species_pvals,
  Significant = ifelse(species_pvals < 0.05, "Yes", "No")
)

# Sort by p-value
species_results <- species_results[order(species_results$P_value),]

print("Species contributing to sampling method differences:")
print(species_results)

# Visualize significant species
sig_species <- species_results[species_results$P_value < 0.05,]

if(nrow(sig_species) > 0) {
  par(mar=c(8,4,2,1))
  barplot(sig_species$Test_Stat, 
          names.arg=sig_species$Species,
          las=2, col="steelblue",
          main="Species with Significant Differences\n(Sampling Method Effect)",
          ylab="Test Statistic")
  abline(h=0, col="red", lty=2)
  par(mar=c(5,4,4,2))
}
```

### mvabund Analysis: Testing Habitat Effects

```{r mvabund_habitat}
# Fit model for habitat
mv_habitat <- manyglm(birds_mv ~ data$Habitat, family="negative.binomial")

# Hypothesis test
mv_habitat_test <- anova(mv_habitat,
                        p.uni="adjusted",
                        nBoot=999,
                        test="LR")

print("Multivariate test results for Habitat:")
print(mv_habitat_test)

pval_habitat <- mv_habitat_test$table["data$Habitat", "Pr(>Dev)"]
if(!is.na(pval_habitat) && pval_habitat < 0.05) {
  # ... rest of code
} else if(!is.na(pval_habitat)) {
  print("✗ No significant difference in community composition by habitat")
} else {
  print("Note: P-value not available - check test results above")
}
```

### Complex Model: Both Factors

```{r mvabund_complex}
# Fit model with both factors
mv_both <- manyglm(birds_mv ~ data$Sample + data$Habitat, family="negative.binomial")

# Test each term
mv_both_test <- anova(mv_both,
                     p.uni="adjusted",
                     nBoot=999,
                     test="LR")

print("Results for model with both factors:")
print(mv_both_test)

# Could also test interaction if sample sizes allow
# mv_interaction <- manyglm(birds_mv ~ data$Sample * data$Habitat, family="negative.binomial")
```

### Visualizing mvabund Results

```{r mvabund_viz, fig.width=12, fig.height=8}
# Plot model residuals
par(mfrow=c(2,2))
plot(mv_sample)
par(mfrow=c(1,1))

# Create abundance plots for significant species
sig_species_names <- species_results$Species[species_results$Significant == "Yes"]

if(length(sig_species_names) > 0) {
  # Select a few species to plot (max 6 for clarity)
  plot_species <- head(sig_species_names, 6)
  
  par(mfrow=c(2,3))
  for(sp in plot_species) {
    boxplot(birds[,sp] ~ data$Sample,
            main=sp,
            xlab="Sampling Method",
            ylab="Abundance",
            col=c("lightblue", "lightcoral", "lightgreen"))
  }
  par(mfrow=c(1,1))
}
```

## Comparing ANOSIM vs. mvabund Approaches

Let's summarize and compare the results from both methods:

```{r comparison}
# Create comparison table
comparison_results <- data.frame(
  Method = c("ANOSIM", "mvabund GLM"),
  Sampling_Method_P = c(
    sampbirds$signif,
    mv_sample_test$table[1,"Pr(>Dev)"]
  ),
  Sampling_Method_Stat = c(
    round(sampbirds$statistic, 3),
    round(mv_sample_test$table[1,"Dev"], 3)
  ),
  Habitat_P = c(
    forestbirds$signif,
    mv_habitat_test$table[1,"Pr(>Dev)"]
  ),
  Habitat_Stat = c(
    round(forestbirds$statistic, 3),
    round(mv_habitat_test$table[1,"Dev"], 3)
  )
)

print("Comparison of Methods:")
print(comparison_results)
```

### When to Use Each Method

**Use ANOSIM/SIMPER when:**
- You need a quick, exploratory analysis
- Your focus is on dissimilarity/distance-based patterns
- You want a simple, non-parametric approach
- Sample sizes are very small
- You're primarily interested in overall community patterns

**Use mvabund when:**
- You want more statistical power
- You need species-level inference with proper p-values
- Your data has count structure with overdispersion
- You want to account for mean-variance relationships
- You need to include covariates or test complex designs
- You want model-based inference rather than distance-based

**Key differences:**

| Aspect | ANOSIM | mvabund |
|--------|--------|---------|
| **Approach** | Distance-based, non-parametric | Model-based GLMs |
| **Test statistic** | R (rank-based) | Likelihood ratio |
| **Species inference** | Indirect (via SIMPER) | Direct with adjusted p-values |
| **Mean-variance** | Not explicitly modeled | Explicitly modeled |
| **Power** | Lower | Higher |
| **Assumptions** | Exchangeability | GLM assumptions |
| **Complexity** | Simple | More complex |

### Methodological Considerations

```{r considerations}
print("Important considerations:")
print("")
print("ANOSIM limitations:")
print("- Sensitive to differences in dispersion (not just location)")
print("- R statistic can be hard to interpret")
print("- Less powerful for detecting subtle differences")
print("- Doesn't provide species-level p-values")
print("")
print("mvabund advantages:")
print("- More powerful statistical tests")
print("- Proper inference at both community and species levels")
print("- Accounts for data properties (counts, overdispersion)")
print("- Can include covariates and interactions")
print("- Model checking via residual diagnostics")
print("")
print("Best practice: Use both methods for complementary insights!")
```

## Final thoughts

ANOSIM and SIMPER are simple tests to test for species composition differences and provide an actual hypothesis test with a proper P value. This approach is more appropriate to analyze species composition data rather than other approaches, such as ANOVA, which may be giving unreliable results.

However, **modern best practice** increasingly favors model-based approaches like mvabund because they:
- Provide greater statistical power
- Allow direct species-level inference
- Better account for properties of count data
- Enable more complex experimental designs
- Provide better control of Type I error rates

### Recommendations for your analyses:

1. **Start with exploratory analysis:** Use ordination (NMDS, PCA) to visualize patterns
2. **Test community-level hypotheses:** Use mvabund for formal testing
3. **Identify key species:** Use species-level tests from mvabund
4. **Validate with ANOSIM:** Use as a complementary non-parametric check
5. **Report both:** When results differ, discuss why (dispersion effects, power differences)

### Additional Resources:

- Wang et al. (2012) "mvabund – an R package for model-based analysis of multivariate abundance data" - Methods in Ecology and Evolution
- Warton et al. (2012) "Distance-based multivariate analyses confound location and dispersion effects" - Methods in Ecology and Evolution
- Warton et al. (2015) "So Many Variables: Joint Modeling in Community Ecology" - Trends in Ecology & Evolution

## Session

```{r session, echo=TRUE}
sessionInfo()
```