---
title: "18 - Biodiversity analysis"
author: "Francisco E. Fonturbel"
date: "21/February/2023"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Using R to conduct biodiversity analysis

Today we will learn how to analyze and make sense of our biodiversity data (after long campaigns in the field collecting data... really exhausting). The `BiodiversityR` package contains many useful functions to run biodiversity analysis (the first time that you load this package may be asked to install XQuartz).


```{r data, message=FALSE, warning=FALSE}
#Loading required packages
library(dplyr)
library(ggplot2)
library(ggsci)
library(ggpubr)
library(vegan)
library(BiodiversityR)
library(Rarefy)

#Loading work data
data = read.csv("data/18_canola.csv",
                   header=TRUE,
                   sep=",")

attach(data)

animal.dt<-data[,5:26]
env.dt<-data[,3:4]

env.dt$posicion<-as.factor(env.dt$posicion)
env.dt$color<-as.factor(env.dt$color)
```

```{r grooming, include=FALSE, warning=FALSE, message=FALSE}
#Enabling multiplot capabilities
source("multiplot.R")

##Some ggplot2 grooming
My_Theme2 = theme(
  axis.title.x = element_text(size = 20,face="bold"),
  axis.text.x = element_text(size = 16),
  axis.text.y = element_text(size = 16),
  axis.title.y = element_text(size = 20,face="bold"),
  legend.text = element_text(size = 14),
  legend.title = element_text(size = 14,face="bold"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_rect(fill = "white"),
  plot.title = element_text(size = 20,face="bold"))
```

Today we will use some data of [Jaques et al. 2023](https://www.mdpi.com/2073-4395/13/2/552) as a working example. This dataset contains data of insect pollinator in a canola crop (the one with the yellow flowers), comparing pan trap color and position effects in the diversity of insects captured.

## Rarefaction curves

How well sampled was our crop? To answer this question we may represent the cumulative number of species in function of the number of samples (pan traps) analyzed:

```{r raref, echo=TRUE, message=FALSE, warning=FALSE}
raref_bb<-rare_alpha(animal.dt, method="hill", random=999, fun_div='speciesdiv', args=a, mean=TRUE)

plot(raref_bb[,1],ylab="Species (cumulative)",xlab="Number of sampling units",type="l", ylim=range(raref_bb,na.rm=TRUE))
lines(raref_bb[,2],lty=2)
lines(raref_bb[,3],lty=2)
```

When we open our first flask, all insects in there are new because we are starting from zero species. Then, we open the second flask we found many new species but some are repeated from the first flask. And if we continue to adding sampling units, the number of new species found decreases. When the rarefaction curve gets **asymptotic** (i.e., no now species are found despite adding more sampling units) we can say that our sampling was complete. In this case, of course, we were unable to capture the whole insect diversity at the canola crop.

So, how can we know how many species are "missing"? Fortunately, many people devoted a lot of time to answer this questions two decades ago and we have many indices of expected species richness, for example:

```{r expS, echo=TRUE, message=FALSE, warning=FALSE}
animal.dt<-data[,5:26]
estimateR(animal.dt)
specpool(animal.dt)
```

Here we have three of the most common indices: Chao1, Jackknife, and Bootstrap. We can simply average them (which is 28.43 species) and compare this expected species richness against the observed species richness (22 species). So, sampling effectiveness is 22 / 28.43 = 0.77, meaning that we sampled about 77% of the insect species present in this crop.

## Rank-abundance curves

Within biological communities is quite common to found very abundant species (the dominant ones) and others that are present in low abundances (also known as rare species). If we have quantitative data, we can estimate a relative abundance of those species and order them from the most abundant to the least:

```{r rank-ab, echo=TRUE, message=FALSE, warning=FALSE}
rab.pol<-rankabundance(animal.dt)
rab.pol

rankabunplot(rab.pol, scale='abundance')
```

As we can see, there are some abundant species that dominate the community, and many others that occur in low abundances.

Let's compare this between pan traps placed at the border or the center of the crop:

**Border**
```{r rank-border}
rab.bor<-rankabundance(animal.dt, y=env.dt, factor="posicion", level="border")
rab.bor

rankabunplot(rab.bor, scale='abundance')
```

**Center**
```{r rank-center}
rab.cen<-rankabundance(animal.dt, y=env.dt, factor="posicion", level="center")
rab.cen

rankabunplot(rab.cen, scale='abundance')
```

We can observe that different insect species dominate in the crop's border or the center.


And now let's compare insect diversity by pan-trap color:

**White**
```{r rank-w}
rab.w<-rankabundance(animal.dt, y=env.dt, factor="color", level="white")
rab.w

rankabunplot(rab.w, scale='abundance')
```

**Yellow**
```{r rank-y}
rab.y<-rankabundance(animal.dt, y=env.dt, factor="color", level="yellow")
rab.y

rankabunplot(rab.y, scale='abundance')
```

**Blue**
```{r rank-b}
rab.b<-rankabundance(animal.dt, y=env.dt, factor="color", level="blue")
rab.b

rankabunplot(rab.b, scale='abundance')
```

Impressive! there are huge differences in dominance. We may go back to lesson 16 and apply an ANOSIM test to this data (or simply read Solange's paper in full).


## Species diversity

Now, we will use the `diversityresult` function to estimate species richness, species diversity (based on the Shannon index, H'), and eveness (based on the Pielou index, J). For each element, we can extact only one comparison level (border vs center, or different colors), and order them into a table below:

```{r diversity, echo=TRUE}
#Spp richness
div_s<-diversityresult(animal.dt, y=NULL, index="richness", method="pooled", digits=3)

##By trap location
div_sb<-diversityresult(animal.dt, y=env.dt, factor="posicion", level="border", index="richness", method="pooled", sortit=TRUE, digits=3)
div_sc<-diversityresult(animal.dt, y=env.dt, factor="posicion", level="center", index="richness", method="pooled", sortit=TRUE, digits=3)

##By colors
div_sw<-diversityresult(animal.dt, y=env.dt, factor="color", level="white", index="richness", method="pooled", sortit=TRUE, digits=3)
div_sy<-diversityresult(animal.dt, y=env.dt, factor="color", level="yellow", index="richness", method="pooled", sortit=TRUE, digits=3)
div_sb<-diversityresult(animal.dt, y=env.dt, factor="color", level="blue", index="richness", method="pooled", sortit=TRUE, digits=3)

#Overall diversity
div_all<-diversityresult(animal.dt, y=NULL, index="Shannon", method="mean", digits=3)

##By trap location
div_border<-diversityresult(animal.dt, y=env.dt, factor="posicion", level="border", index="Shannon", method="mean", sortit=TRUE, digits=3)
div_center<-diversityresult(animal.dt, y=env.dt, factor="posicion", level="center", index="Shannon", method="mean", sortit=TRUE, digits=3)

##By colors
div_w<-diversityresult(animal.dt, y=env.dt, factor="color", level="white", index="Shannon", method="mean", sortit=TRUE, digits=3)
div_y<-diversityresult(animal.dt, y=env.dt, factor="color", level="yellow", index="Shannon", method="mean", sortit=TRUE, digits=3)
div_b<-diversityresult(animal.dt, y=env.dt, factor="color", level="blue", index="Shannon", method="mean", sortit=TRUE, digits=3)

#Evenness
div_ev<-diversityresult(animal.dt, y=NULL, index="Eevenness", method="mean", digits=3)

##By trap location
div_eb<-diversityresult(animal.dt, y=env.dt, factor="posicion", level="border", index="Eevenness", method="mean", sortit=TRUE, digits=3)
div_ec<-diversityresult(animal.dt, y=env.dt, factor="posicion", level="center", index="Eevenness", method="mean", sortit=TRUE, digits=3)

##By colors
div_ew<-diversityresult(animal.dt, y=env.dt, factor="color", level="white", index="Eevenness", method="mean", sortit=TRUE, digits=3)
div_ey<-diversityresult(animal.dt, y=env.dt, factor="color", level="yellow", index="Eevenness", method="mean", sortit=TRUE, digits=3)
div_eb<-diversityresult(animal.dt, y=env.dt, factor="color", level="blue", index="Eevenness", method="mean", sortit=TRUE, digits=3)
```

#### Species richness

| Comparison | S |
| ---------- | -- |
| Overall | `r div_s` |
|  |  |
| Border | `r div_sb` |
| Center | `r div_sc` |
|  |  |
| White | `r div_sw` |
| Yellow | `r div_sy` |
| Blue | `r div_sb` |

#### Diversity


| Comparison | H' |
| ---------- | -- |
| Overall | `r div_all` |
|  |  |
| Border | `r div_border` |
| Center | `r div_center` |
|  |  |
| White | `r div_w` |
| Yellow | `r div_y` |
| Blue | `r div_b` |


#### Evenness

| Comparison | E |
| ---------- | -- |
| Overall | `r div_ev` |
|  |  |
| Border | `r div_eb` |
| Center | `r div_ec` |
|  |  |
| White | `r div_ew` |
| Yellow | `r div_ey` |
| Blue | `r div_eb` |


While we can calculate this by hand, with a pencil, a piece of papers, and a simple calculator (as I did 20 years ago when was an undergraduate student), this procedure simplifies such calculations, particularly when we have too many species.


## Final thoughts

Biodiversity analysis in R is simple and straightforward. Here I presented only a few options, but I highly encourage you to explore the `BiodiversityR` package a little bit further on your own.


## Session

```{r session, echo=TRUE}
sessionInfo()
```