---
title: "19 - Time series analysis"
author: "Francisco E. Fonturbel"
date: "21/February/2023"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## A very basic introduction to Time series analysis (TSA)

Today we will learn how to analyze and make sense of our biodiversity data (after long campaigns in the field collecting data... really exhausting). The `BiodiversityR` package contains many useful functions to run biodiversity analysis (the first time that you load this package may be asked to install XQuartz).


```{r data, message=FALSE, warning=FALSE}
#Loading required packages
library(dplyr)
library(ggplot2)
library(ggsci)
library(ggpubr)
library(vegan)
library(BiodiversityR)
library(Rarefy)

#Loading work data
data = read.csv("data/18_canola.csv",
                   header=TRUE,
                   sep=",")

attach(data)

animal.dt<-data[,5:26]
env.dt<-data[,3:4]

env.dt$posicion<-as.factor(env.dt$posicion)
env.dt$color<-as.factor(env.dt$color)
```




## Final thoughts

Biodiversity analysis in R is simple and straightforward. Here I presented only a few options, but I highly encourage you to explore the `BiodiversityR` package a little bit further on your own.


## Session

```{r session, echo=TRUE}
sessionInfo()
```