---
title: "19 - Time series analysis"
author: "Francisco E. Fonturbel"
date: "21/February/2023"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## A very basic introduction to Time series analysis (TSA)

When we measure the same variable (whichever it is) many times in different times (e.g., daily temperature, monthly moth abundance, yearly inflation) we can build a times series dataset, allowing to conduct very interesting analyses. With TSA we can assess trends without the noise of temporal variation (e.g., seasons) in our data, and therefore making more robust inferences.


```{r data, message=FALSE, warning=FALSE}
#Loading required packages
library(TSA)
library(forecast)
library(astsa)

#Loading work data
data<-read.table("data/19_bird_data.txt", header=T)
attach(data)
names(data)
```

We will use now an unpublished dataset of bird abundance in two habitat types (notro and canelo forests) with monthly records for 12 year (yeah, a lot of work).

First, we need to convert each variable into a TSA object:

```{r objects}
aphcan<-ts(aphspiC, start=c(2003, 1), end=c(2015, 12), frequency=12)
aphnot<-ts(aphspiN, start=c(2003, 1), end=c(2015, 12), frequency=12)
carcan<-ts(carbarC, start=c(2003, 1), end=c(2015, 12), frequency=12)
carnot<-ts(carbarN, start=c(2003, 1), end=c(2015, 12), frequency=12)
elacan<-ts(elaalbC, start=c(2003, 1), end=c(2015, 12), frequency=12)
elanot<-ts(elaalbN, start=c(2003, 1), end=c(2015, 12), frequency=12)
phrcan<-ts(phrpatC, start=c(2003, 1), end=c(2015, 12), frequency=12)
phrnot<-ts(phrpatN, start=c(2003, 1), end=c(2015, 12), frequency=12)
trocan<-ts(troaedC, start=c(2003, 1), end=c(2015, 12), frequency=12)
tronot<-ts(troaedN, start=c(2003, 1), end=c(2015, 12), frequency=12)
turcan<-ts(turfalC, start=c(2003, 1), end=c(2015, 12), frequency=12)
turnot<-ts(turfalN, start=c(2003, 1), end=c(2015, 12), frequency=12)
zoncan<-ts(zoncapC, start=c(2003, 1), end=c(2015, 12), frequency=12)
zonnot<-ts(zoncapN, start=c(2003, 1), end=c(2015, 12), frequency=12)
```

Now that we have those object, we can plot each variable:

```{r plots_hab}
plot(aphcan,ylab='Rayadito abundance (Canelo)')
plot(aphnot,ylab='Rayadito abundance (Notro)')
plot(carcan,ylab='Jilguero abundance (Canelo)')
plot(carnot,ylab='Jilguero abundance (Notro)')
plot(elacan,ylab='Fiofio abundance (Canelo)')
plot(elanot,ylab='Fiofio abundance (Notro)')
plot(phrcan,ylab='Cometocino abundance (Canelo)')
plot(phrnot,ylab='Cometocino abundance (Notro)')
plot(trocan,ylab='Chercan abundance (Canelo)')
plot(tronot,ylab='Chercan abundance (Notro)')
plot(turcan,ylab='Zorzal abundance (Canelo)')
plot(turnot,ylab='Zorzal abundance (Notro)')
plot(zoncan,ylab='Chincol abundance (Canelo)')
plot(zonnot,ylab='Chincol abundance (Notro)')
```

Holy Molly! a lot of data! Let's unify habitat data by species:

```{r unify}
aphspi<-cbind(aphspiC,aphspiN)
carbar<-cbind(carbarC,carbarN)
elaalb<-cbind(elaalbC,elaalbN)
phrpat<-cbind(phrpatC,phrpatN)
troaed<-cbind(troaedC,troaedN)
turfal<-cbind(turfalC,turfalN)
zoncap<-cbind(zoncapC,zoncapN)
```

And convert them in TSA objects:

```{r tsa_unif}
rayadito<-ts(aphspi, start=c(2003, 1), end=c(2015, 12), frequency=12)
jilguero<-ts(carbar, start=c(2003, 1), end=c(2015, 12), frequency=12)
fiofio<-ts(elaalb, start=c(2003, 1), end=c(2015, 12), frequency=12)
cometocino<-ts(phrpat, start=c(2003, 1), end=c(2015, 12), frequency=12)
chercan<-ts(troaed, start=c(2003, 1), end=c(2015, 12), frequency=12)
zorzal<-ts(turfal, start=c(2003, 1), end=c(2015, 12), frequency=12)
chincol<-ts(zoncap, start=c(2003, 1), end=c(2015, 12), frequency=12)
```

And plot them again:

```{r plot_unif}
plot(rayadito, main="Aphrastura spinicauda")
plot(jilguero, main="Carduelis barbata")
plot(fiofio, main="Elaenia albiceps")
plot(cometocino, main="Phrygilus patagonicus")
plot(chercan, main="Troglodytes aedon")
plot(zorzal, main="Turdus falcklandii")
plot(chincol, main="Zonotrichia capensis")
```


## Autocorrelation functions

As we dealt with spatial autocorrelation on lesson 15, in TSA we will deal with temporal autocorrelation, that has the same principle (the shorter the measurement interval, the more similar a variable would be). In this case, we will estimate the autocorrelation function (ACF) for each variable:

```{r acf}
acf(as.vector(aphcan),lag.max=36, main="ACF Rayadito Canelo")
acf(as.vector(aphnot),lag.max=36, main="ACF Rayadito Notro")
acf(as.vector(carcan),lag.max=36, main="ACF Jilguero Canelo")
acf(as.vector(carnot),lag.max=36, main="ACF Jilguero Notro")
acf(as.vector(elacan),lag.max=36, main="ACF Fiofio Canelo")
acf(as.vector(elanot),lag.max=36, main="ACF Fiofio Notro")
acf(as.vector(phrcan),lag.max=36, main="ACF Cometocino Canelo")
acf(as.vector(phrnot),lag.max=36, main="ACF Cometocino Notro")
acf(as.vector(trocan),lag.max=36, main="ACF Chercan Canelo")
acf(as.vector(tronot),lag.max=36, main="ACF Chercan Notro")
acf(as.vector(turcan),lag.max=36, main="ACF Zorzal Canelo")
acf(as.vector(turnot),lag.max=36, main="ACF Zorzal Notro")
acf(as.vector(zoncan),lag.max=36, main="ACF Chincol Canelo")
acf(as.vector(zonnot),lag.max=36, main="ACF Chincol Notro")
```

It's common to find regular "waves" of autocorrelation, as they are given by seasonality. This analysis is useful to understand the temporal variation in any phenomenon that we are studying!


## Decomposing time series data

This is my favorite part! We can decompose a time series data into four elements: the data itself, the seasonal variation, the trend, and the remainder (error). We also can plot the variation within each month across years, and annual variation across sampling years. We will use only two species (one resident and one migratory) as an example here:

```{r decomp}
#Rayadito-Canelo
decompose(aphcan)
fit.aphcan<-stl(aphcan, "periodic")
plot(fit.aphcan, main="Rayadito-Canelo")
monthplot(aphcan, main="Rayadito-Canelo")
seasonplot(aphcan, main="Rayadito-Canelo")
##
#Rayadito-Notro
decompose(aphnot)
fit.aphnot<-stl(aphnot, "periodic")
plot(fit.aphnot, main="Rayadito-Notro")
monthplot(aphnot, main="Rayadito-Notro")
seasonplot(aphnot, main="Rayadito-Notro")

##
#Chincol-Canelo
decompose(zoncan)
fit.zoncan<-stl(zoncan, "periodic")
plot(fit.zoncan, main="Chincol-Canelo")
monthplot(zoncan, main="Chincol-Canelo")
seasonplot(zoncan, main="Chincol-Canelo")
##
#Chincol-Notro
decompose(zonnot)
fit.zonnot<-stl(zonnot, "periodic")
plot(fit.zonnot, main="Chincol-Notro")
monthplot(zonnot, main="Chincol-Notro")
seasonplot(zonnot, main="Chincol-Notro")
```

Very useful information indeed, with a bunch of applications for population ecology, conservation biology, wildlife management, etc.

### Cross-correlation functions

The cross-correlation functions (CCF) allows us to examine the variation of one variable in function of other. In the case of the example, how the abundances of two bird species vary in time or the same species in two habitats, let's take a look on a couple of examples:

```{r ccf}
ccf(aphcan, aphnot)
var.test(aphcan, aphnot)

ccf(zoncan, zonnot)
var.test(zoncan, zonnot)
```


## ARIMA models

Autoregressive integrated moving average (aka ARIMA) are intended to be used with time series data as they account for seasonality and temporal autocorrelation. Below we will compare one species in two habitats as an example using ARIMA models of first order. ARIMA could be decomposed in AR-I-MA and 0 or 1 indicate if each component (autoregressive, integrated, or moving average) is present or not in the model. A detailed explanation of these models is beyond the scope of this lesson, but there are many good books on this subject.

```{r arima}
##Rayadito-Canelo
m1.aphcan=arima(aphcan,order=c(1,1,1),seasonal=list(order=c(0,1,1), period=12))
m1.aphcan
#Diagnosing ARIMA model
tsdiag(m1.aphcan)
Box.test(m1.aphcan$residuals,lag=1)
#Estimating ma1 significance
(1-pnorm(abs(m1.aphcan$coef)/sqrt(diag(m1.aphcan$var.coef))))*2
# Residuals
plot(window(rstandard(m1.aphcan),start=c(2003,2)), ylab='Standardized Residuals',type='o')
abline(h=0)
# ACF
acf(as.vector(window(rstandard(m1.aphcan),start=c(2003,2))), lag.max=36)
hist(window(rstandard(m1.aphcan),start=c(2003,2)), xlab='Standardized Residuals')


##Rayadito-Notro
m1.aphnot=arima(aphnot,order=c(1,1,1),seasonal=list(order=c(0,1,1), period=12))
m1.aphnot
#Diagnosing ARIMA model
tsdiag(m1.aphnot)
Box.test(m1.aphnot$residuals,lag=1)
#Estimating ma1 significance
(1-pnorm(abs(m1.aphnot$coef)/sqrt(diag(m1.aphnot$var.coef))))*2
# Residuals
plot(window(rstandard(m1.aphnot),start=c(2003,2)), ylab='Standardized Residuals',type='o')
abline(h=0)
# ACF
acf(as.vector(window(rstandard(m1.aphnot),start=c(2003,2))), lag.max=36)
hist(window(rstandard(m1.aphnot),start=c(2003,2)), xlab='Standardized Residuals')
```

The output of those models is a little difficult to interpret at first, but after some reading and practice everything makes sense.


## Final thoughts

Time Series Analysis is a powerful tool to work with measurements repeated in time.


## Session

```{r session, echo=TRUE}
sessionInfo()
```