---
title: "19 - Time series analysis"
author: "Francisco E. Fonturbel"
date: "21/February/2023"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## A very basic introduction to Time series analysis (TSA)

When we measure the same variable (whichever it is) many times in different times (e.g., daily temperature, monthly moth abundance, yearly inflation) we can build a times series dataset, allowing to conduct very interesting analyses. With TSA we can assess trends without the noise of temporal variation (e.g., seasons) in our data, and therefore making more robust inferences.


```{r data, message=FALSE, warning=FALSE}
#Loading required packages
library(TSA)
library(forecast)
library(astsa)

#Loading work data
data<-read.table("data/19_bird_data.txt", header=T)
attach(data)
names(data)
```

We will use now an unpublished dataset of bird abundance in two habitat types (notro and canelo forests) with monthly records for 12 year (yeah, a lot of work).

First, we need to convert each variable into a TSA object:

```{r objects}
aphcan<-ts(aphspiC, start=c(2003, 1), end=c(2015, 12), frequency=12)
aphnot<-ts(aphspiN, start=c(2003, 1), end=c(2015, 12), frequency=12)
carcan<-ts(carbarC, start=c(2003, 1), end=c(2015, 12), frequency=12)
carnot<-ts(carbarN, start=c(2003, 1), end=c(2015, 12), frequency=12)
elacan<-ts(elaalbC, start=c(2003, 1), end=c(2015, 12), frequency=12)
elanot<-ts(elaalbN, start=c(2003, 1), end=c(2015, 12), frequency=12)
phrcan<-ts(phrpatC, start=c(2003, 1), end=c(2015, 12), frequency=12)
phrnot<-ts(phrpatN, start=c(2003, 1), end=c(2015, 12), frequency=12)
trocan<-ts(troaedC, start=c(2003, 1), end=c(2015, 12), frequency=12)
tronot<-ts(troaedN, start=c(2003, 1), end=c(2015, 12), frequency=12)
turcan<-ts(turfalC, start=c(2003, 1), end=c(2015, 12), frequency=12)
turnot<-ts(turfalN, start=c(2003, 1), end=c(2015, 12), frequency=12)
zoncan<-ts(zoncapC, start=c(2003, 1), end=c(2015, 12), frequency=12)
zonnot<-ts(zoncapN, start=c(2003, 1), end=c(2015, 12), frequency=12)
```

Now that we have those object, we can plot each variable:

```{r plots_hab}
plot(aphcan,ylab='Rayadito abundance (Canelo)')
plot(aphnot,ylab='Rayadito abundance (Notro)')
plot(carcan,ylab='Jilguero abundance (Canelo)')
plot(carnot,ylab='Jilguero abundance (Notro)')
plot(elacan,ylab='Fiofio abundance (Canelo)')
plot(elanot,ylab='Fiofio abundance (Notro)')
plot(phrcan,ylab='Cometocino abundance (Canelo)')
plot(phrnot,ylab='Cometocino abundance (Notro)')
plot(trocan,ylab='Chercan abundance (Canelo)')
plot(tronot,ylab='Chercan abundance (Notro)')
plot(turcan,ylab='Zorzal abundance (Canelo)')
plot(turnot,ylab='Zorzal abundance (Notro)')
plot(zoncan,ylab='Chincol abundance (Canelo)')
plot(zonnot,ylab='Chincol abundance (Notro)')
```


## Final thoughts

Biodiversity analysis in R is simple and straightforward. Here I presented only a few options, but I highly encourage you to explore the `BiodiversityR` package a little bit further on your own.


## Session

```{r session, echo=TRUE}
sessionInfo()
```