---
title: "20 - Circular statistics and daily activity patterns"
author: "Francisco E. Fonturbel"
date: "01/November/2025"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Circular statistics: a matter of time

When we consider time as a variable we face a little big problem: the beginning and the end points are the same. When we run from point A to point B, we get farther from the start point and near to the end point (let's call it "linear data"). But, what if we run around a tree or a fountain? in this case the finishing point will be the same in which we started (let's call it "circular data"). the same happens with time... 00:01 h and 23:59 h seems far away but are quite close indeed (only two minutes away). So, using data analysis approaches intended for linear data to analyze circular data is, of course, a gross mistake. And that is why circular statistics exist.


```{r data, message=FALSE, warning=FALSE}
#Loading required packages
library("circular")
library("plotrix")

#Loading an example dataset
activ<-read.table("data/20_activity.txt", header= T)
```

Here I'll reuse part of the script and data of a public repository co-authored with Javier Cuadra (a former student). If you want to take a look on the complete material that we prepared, just click [HERE](https://github.com/fonturbel-lab/AnimActivity).

In this example dataset we have a lot of activity data of the hummingbird _Sephanoides sephaniodes_ gathered using camera traps for Javier's dissertation (the paper is available [here](https://link.springer.com/article/10.1007/s10530-021-02519-2)).

Before we start, we need to convert our hour data into radians using the following formula:

$$
rad=time*2*{\pi}
$$
**Important:** Prior to transform time data to radians you must convert your time figures to a number.

```{r radian, echo=TRUE}
radian <- activ$Time_2*2*pi
data <- radian[activ$Condition == "A"]
```

Then, we have to convert our radian data into a `circular class` object in order to be used with the `circular` package.

```{r circ_class, echo=TRUE, warning=FALSE}

as.circular(data)
n_a <- circular(data%%24, 
                     units="radians", template="clock24")

```


## Rose diagrams

A simple representation of activity data is the rose diagram, which is obtained using the function `rose.diag` to represent a circular dataset into a 24h clock.

```{r rose_diag, echo=TRUE, eval=TRUE, fig.width=8}
rose.diag(n_a, bin = 24, col = "lightblue", main = "Rose diagram", 
          prop = 2)
```

As we can see, hummingbird activity is diurnal and concentrated in a relatively narrow period of the day.


## Frequency plots

A more elaborated representation of activity periods than the rose diagram is the frequency plot. On a frequency plot, the lenght of each bar is proportional to .

To make a decent frequency plot we need to arrange our data in a table with frequency of records per hour. We'll load a new file (ready to go) as example.

Then, we will use the `clock24.plot` function of the `plotrix` package.

```{r freq_plot, echo=TRUE, eval=TRUE, fig.width=8}
#Loading new data file
activ_table<-read.table("data/20_records_hour.txt", header= T)

#Making the plot
clock<-c(0:23) 
clock24.plot(activ_table$A,clock, minutes=T, label.prop=c(1.15,1.15),show.grid.labels = 3,  show.grid=T, grid.bg="white", grid.left=T,grid.col="black",rad.col="black", rp.type="r",lwd=5,line.col="blue",lty=1, main="",add=F)
```

Another way to represent the frequency plot is with the magnitudes represented by the positions of the vertices of the polygon. In this plot, the positions of the vertices indicate the frequency of the records.

```{r freq_plot2, echo=TRUE, eval=TRUE, fig.width=8}

#Making the plot
clock<-c(0:23) 
mi.azul <- rgb(0, 0, 1, 0.6) # "#0000FF99"
clock24.plot(activ_table$A,clock,minutes=T,label.prop=c(1.15,1.15),show.grid.labels = 3, rp.type="s",point.symbols=16,point.col="#0000FF99",add=F)
clock24.plot(activ_table$A,clock, rp.type="p",poly.col="#0000FF99",lwd=4,line.col="#0000FF99",lty=1, main="",add=T) 


```

Very impressive, right?


## Formal tests

Now that we have made such beautiful rose diagrams and frequency plots, we maybe will be interested into test some hypotheses about activity patterns. The most simple question here will be if animal activity is homogeneous or not along the day. Of course our _null hypothesis_ will be that animal activity is uniform along the day (like a faculty researcher, for example). The `circular` package has two common tests to do that: Rao's test fpr homogeneity of angular data (`rao.test` function), and Rayleight test of Unifortmity (`rayleigh.test` function). Both tests do basically the same, the difference is that Rayleigh test needs only one `circular.class` element whereas Rao's test need two samples to be compared.

Let's see how they work:

```{r rr_tests, echo=TRUE, eval=TRUE, fig.width=8}
#Rao's test
x <- rvonmises(100, circular(0), kappa=10)
y <- rvonmises(100, circular(0), kappa=10)

rao.test(x, y)

#Rayleigh's test
rayleigh.test(n_a)
```



## Daily activity patterns

Another interesting area that we can explore in circular statistics are daily activity patterns. The time of the day is circular, of course, and animals tend to keep a regular activity peaks along the day, with the exception of a few species (e.g., foxes) that can show activity throughout the day. In this case, we can (1) use kernel estimators to depict daily activity patterns along a 24-h period, and (2) use the `compareCkern` function from the `activity` package to compare two kernels. Let's reuse Javier's data once again:


```{r data_activk, echo=TRUE}
#Loading required packages
library("overlap")
library ("activity")
#Loading two activity datasets (time)
act_x<-read.table("./data/20_act_x.txt", header= T)
act_y<-read.table("./data/20_act_y.txt", header= T)
```


Before plotting the kernels, we must transform time data to radians using the following formula:

$$
rad=time*2*{\pi}
$$

```{r rad, echo=TRUE, eval=TRUE}
rad_x<-act_x$Time_2*2*pi
rad_y<-act_y$Time_2*2*pi
```

And one more step before the fun start... we must compute fitted data. In this example, we will use only one condition (A) of the two conditions available in our datasets A and B.

```{r fit, echo=TRUE, eval=TRUE, warning=FALSE, error=FALSE}
fitact_a <-fitact(rad_x[act_x$Condition=="A"])
fitact_b <-fitact(rad_y[act_y$Condition=="A"])
```

### Density plots

Now we will plot kernel density functions for our two datasets separately.

```{r denplot, echo=TRUE, eval=TRUE, fig.width=8}
#Density plot for dataset A
plot(fitact_a, yunit="dens", las=1, dline=list(col="lightblue"), tline=list(col="blue"), main="Data X", ylim=c(0,0.17))

#Density plot for dataset B
plot(fitact_b, yunit="dens", las=1, dline=list(col="pink"), tline=list(col="red"), main = "Data Y")
```

And we also are able to depict both kernels in one plot:

```{r denpcomp, echo=TRUE, eval=TRUE, fig.width=8}
plot(fitact_a, yunit="dens", las=1, dline=list(col="lightblue"), tline=list(col="blue"), ylim=c(0,0.17))
     plot(fitact_b, yunit="dens", add=TRUE, dline=list(col="pink"), tline=list(col="red"))
     legend("topright", inset = c(0,0), title = "Data", c("X", "Y"), lty= c(1,1), col = c("lightblue", "pink"), bty = "n")
```

It is very nice, isn't it?

## Overlap coefficient _dhat5_

The overlap coefficient of is the area under a density curve.This index is merely descriptive. On the other hand, the function `CompareCkern` of the package `activity`, is a probability test that assess if two datasets come from the same activity distribution. In addition, this test calculate overlap index _dhat5_ for two fitted distributions (with function `fitact`).

Below we compare datasets X and Y, and test if they come or not from the same activity distribution:

```{r comp, echo=TRUE}
compareCkern(fitact_a, fitact_b, reps = 100)
```

## Overlap assessment

Before plotting overlaps we must extracti data for condition = A...

```{r extr, echo=TRUE}
act_a <- rad_x[act_x$Condition=="A"]
act_b <- rad_y[act_y$Condition=="A"]
```

And now we will graph the Density Plots!

```{r densplot, echo=TRUE, eval=TRUE, fig.width=8}
#Density plot for A
densityPlot(act_a, xscale = 24, xcenter = c("noon", "midnight"), add = F, rug = T, extend = "lightgray", main = "Data X", xlab = "Time", ylab = "Density", bty = "L")

#Density plot for B
densityPlot(act_b, xscale = 24, xcenter = c("noon", "midnight"), add = F, rug = T, extend = "lightgray", main = "Data Y", xlab = "Time", ylab = "Density", bty = "L")
```

And of course, we can plot the two datasets in same plot with `overlap`

```{r dplotcomb, echo=TRUE, eval=TRUE, fig.width=8}
overlapPlot(act_a, act_b, main = " ", linecol = c("black", "darkgreen"), lty = c(1,3), bty = "L", xlab = "Time", ylab = "Density")
legend("topright", inset = c(0,0), title = "Data", c("X", "Y"), lty= c(1,3), col = c("gray25", "seagreen"), bty = "n")
```


## Final thoughts

Circular statistics are a different way to assess temporal data. As you may see in the example, it would be absurd to use ANOVA tests or GLM models in this case. What we need is an appropriate tool that takes time circularity into account. Of course there are much more in circular statistics than I presented here, and fortunately there are many good books that you can read if you want to explore this subject further.


## Session

```{r session, echo=TRUE}
sessionInfo()
```