---
title: "20 - Circular statistics"
author: "Francisco E. Fonturbel"
date: "21/February/2023"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## A very basic introduction to Time series analysis (TSA)

When we measure the same variable (whichever it is) many times in different times (e.g., daily temperature, monthly moth abundance, yearly inflation) we can build a times series dataset, allowing to conduct very interesting analyses. With TSA we can assess trends without the noise of temporal variation (e.g., seasons) in our data, and therefore making more robust inferences.


```{r data, message=FALSE, warning=FALSE}
#Loading required packages
library(TSA)
library(forecast)
library(astsa)

#Loading work data
data<-read.table("data/19_bird_data.txt", header=T)
attach(data)
names(data)
```




## Final thoughts

aaa


## Session

```{r session, echo=TRUE}
sessionInfo()
```