---
title: "20 - Circular statistics"
author: "Francisco E. Fonturbel"
date: "21/February/2023"
output: 
  html_document:
    includes:
        after_body: footer.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Circular statistics: a matter of time

When we consider time as a variable we face a little big problem: the beginning and the end points are the same. When we run from point A to point B, we get farther from the start point and near to the end point (let's call it "linear data"). But, what if we run around a tree or a fountain? in this case the finishing point will be the same in which we started (let's call it "circular data"). the same happens with time... 00:01 h and 23:59 h seems far away but are quite close indeed (only two minutes away). So, using data analysis approaches intended for linear data to analyze circular data is, of course, a gross mistake. And that is why circular statistics exist.


```{r data, message=FALSE, warning=FALSE}
#Loading required packages
library("circular")
library("plotrix")

#Loading an example dataset
activ<-read.table("data/20_activity.txt", header= T)
```

Here I'll reuse part of the script and data of a public repository co-authored with Javier Cuadra (a former student). If you want to take a look on the complete material that we prepared, just click [HERE](https://github.com/fonturbel-lab/AnimActivity).


## Final thoughts

aaa


## Session

```{r session, echo=TRUE}
sessionInfo()
```