---
title: "Best Practices for Managing and Analyzing Biological Data"
author: "Francisco E. Fontúrbel"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# Packages used in this document
# install.packages(c("dplyr", "stringr", "knitr", "janitor", "lubridate", "kableExtra"))
library(dplyr)
library(stringr)
library(knitr)
library(janitor)
library(lubridate)
library(kableExtra)
```

# Purpose of this document

This document brings together a set of best practices for data management and analysis,
organized as a reference protocol for lab work and for anyone doing a thesis or graduate
coursework. The idea is simple: a dataset or a script should be findable, openable, understandable,
and reusable by anyone (including yourself, a year later), without depending on the person who
created it being available to explain it.

This lesson is directly tied to the **FAIR** principles (*Findable, Accessible, Interoperable,
Reusable*), which we use as the connecting thread of the whole document rather than treating them
as a box to check for an editorial policy.

- **Findable**: unique, consistent identifiers (e.g., DOIs), complete metadata, file and object
  names that can be understood without opening the file.
- **Accessible**: open formats (plain text, CSV) rather than formats tied to a specific software
  license; raw data backed up and available.
- **Interoperable**: recognized standards (e.g., Darwin Core for biodiversity data), controlled
  vocabularies instead of free-text typing, consistent language between code and publication.
- **Reusable**: version-controlled and commented code, a reproducible project structure, and
  enough documentation for someone else to repeat the analysis without having to guess.

As an illustration of what happens when one of these four pillars fails, Section 7 summarizes a
real case of a biodiversity data audit; the rest of the document (Sections 2 through 6) is the
general protocol that applies to any lab dataset or analysis, whether or not a case like that one
is involved.

---

# Data management best practices

## Why Excel is not enough

- Extra spaces, commas and decimal points, invisible line breaks, and mixed formats
  (text/number/date in the same column) are not always visible when opening the file in Excel,
  but they do break analysis pipelines.
- Under a Spanish-language regional setting, Excel uses `,` as the decimal separator and `.` as
  the thousands separator (or vice versa, depending on the configuration), which produces silent
  errors when exporting to CSV or when opening the file on another computer with a different
  configuration. Swapping a period for a comma (or vice versa) can ruin an entire spreadsheet
  without it being noticeable at a glance.
- **Lab rule**: Excel is used only to **view** data, never as a cleaning or final data-wrangling
  tool.
- To load data into R, always use `.txt` or `.csv`, never read the Excel file directly: the
  libraries that do that (`readxl` and similar) tend to import the same formatting errors carried
  over from the original file.

## Recommended data-cleaning tools

```{r tools-table, echo=FALSE}
tools_table <- tribble(
  ~Tool, ~`Main use`, ~`When to use it`,
  "Notepad++", "View and edit the file as plain text; detect invisible characters, line breaks, encoding", "Before importing any CSV into R, especially if it comes from Excel or from copy/pasting from Word/PDF",
  "OpenRefine", "Bulk data cleaning: text clustering (groups spelling variants of the same value, e.g. 'Shangri-La' / 'Shangrila' / 'Shangri-la'), faceted transformation, reversible change history", "Before any analysis, especially in categorical fields with free-text entry (localities, collector names, taxonomy)"
)

kable(tools_table, caption = "Data-cleaning tools") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

Running the data through Notepad++ or OpenRefine before analyzing it is the step that catches -in
seconds- inconsistencies that would otherwise only show up as strange results halfway through the
analysis.

## Golden rule: the original spreadsheet is never touched

- The raw data file is saved **read-only**, with a name and date, and is never edited manually.
- All cleaning is done through **version-controlled code** that reads the raw file and writes a
  processed file (`data/raw/` → `data/processed/`).
- This preserves traceability: anyone can rebuild the processed dataset from scratch and audit
  exactly which transformation was applied — something impossible if the original Excel file is
  edited by hand and leaves no trace.

## Backing up your data

A "read-only" raw file is useless if it lives as a single copy on a disk that can fail. Backup is
the other half of the golden rule:

- **The 3-2-1 rule** as a reasonable minimum: at least **3** copies of any important dataset, on
  **2** different media (e.g., local disk and cloud), with **1** copy off the everyday working
  machine (Google Drive, GitHub, an external disk that doesn't travel with the computer).
- Raw field data (spreadsheets, camera-trap photos, GPX files) is backed up **the same day** as
  the field trip, before the original file ends up as the only copy on an SD card or a phone.
- A Git repository backs up code and its history, but it **does not replace** backup of heavy raw
  data (photos, rasters, files larger than a couple of MB): for that, use Drive, Dropbox, or
  another cloud storage service.
- Before any major cleaning or restructuring of a dataset, it is worth confirming that the
  previous version is still available in some backup, not only in the "undo" history of whatever
  editor is being used.

---

# Analysis and coding best practices

- **If the script starts with `setwd()` and an absolute path, something is wrong.** Use R Projects
  (`.Rproj`) instead, so that paths are always relative to the project: when sharing the folder
  (or cloning the repo), nobody has to rewrite paths that only exist on one computer.
- Create sub-folders within the project to organize the different files (`data`, `R`, `scripts`,
  `output`, etc.), as in the layout shown below.

## Suggested R project structure

```
project_name/
├── project_name.Rproj         # use R Projects: relative paths, full portability
├── data/
│   ├── raw/                   # NEVER edited by hand; read-only
│   └── processed/              # generated 100% by code
├── R/                          # custom functions (see Section 5)
├── scripts/                    # analysis scripts, numbered in execution order
│   ├── 01_import.R
│   ├── 02_clean.R
│   └── 03_analysis.R
├── output/
│   ├── figures/
│   └── tables/
├── docs/                        # rendered .Rmd reports
├── README.md
└── .gitignore
```

## File and object names

A file or R object name has to be readable and understandable without opening the file or running
the script. Two equally problematic extremes:

- **Too short** (`d1.csv`, `df2`, `m3`): six months later, nobody remembers what it contained or
  which model it represented.
- **Too long** (`final_definitive_data_v3_corrected_use_this_march22.csv`): besides being
  cumbersome to type, it is usually a sign that there is no real version-control system in place
  (that's what Git is for, not the file name). **NOTE:** on large hard drives with many files,
  these long names increase the likelihood of causing data corruption (and loss).

Practical lab conventions:

- Short, descriptive names, in lowercase, with underscores as separators (`snake_case`): e.g.,
  `nahuelbuta_birds_2024.csv` instead of `Nahuelbuta BIRDS Data (2024) FINAL.csv`.
- No spaces, accents, or special characters in file names (spaces cause problems when calling
  paths from the terminal or from other operating systems).
- Scripts numbered according to the order in which they run (`01_import.R`, `02_clean.R`,
  `03_analysis.R`), as in the project structure above.
- Object names within the script follow the same logic: `fit_glmm_fruitset` is more useful than
  `model3` or `glmm_negative_binomial_model_with_random_effects_by_site_final`.
- Versioning lives in Git (commits with descriptive messages), not in the file name (`_v1`,
  `_v2_final`, `_final_final`). If a file needs a version suffix to be understood, that's a sign
  that version control is missing, not that a better suffix is needed.

## Git / GitHub

- **A real change history**: each `commit` is logged with author, date, and the exact diff (which
  line changed, by whom, and when) — something a spreadsheet shared by email never offers.
- **Basic workflow**: `git pull` (fetch changes) → edit → `git add` → `git commit -m "descriptive
  message"` → `git push` (upload changes).
- **Branches**: allow you to "experiment" with code or analysis without touching the main version
  (`main`); if the experiment doesn't work out, it can be discarded without leaving any trace on
  the main branch.
- **Fork**: lets you start from another person's or lab's repository and develop it
  independently, while keeping the connection to the original.
- **RStudio integration**: RStudio has native Git integration (the "Git" panel: commit, push,
  pull, history) without leaving the IDE. Authentication in RStudio is set up once and stays
  connected from then on. It's also useful to install the `GitHub Desktop` app.

## Commented scripts and R Markdown

- Use clearly commented scripts that explain the *why* of each step, not just the *what* the line
  of code does.
- Even better is to use R Markdown files (`.Rmd`), which combine narrative text with executable
  code (R, and also Python, bash, etc.), so that the final document (HTML/PDF/Word) is generated
  automatically from the actual data, avoiding manual copy-pasting of results — another classic
  source of transcription errors.
- Markdown (`.md`) is plain text, lightweight, and readable on any platform, without depending on
  proprietary software (unlike Word's `.docx`). As an additional reference: language models
  process Markdown far more efficiently than binary formats such as `.docx` or `.pdf`, which also
  matters if you plan to use AI as support for analysis or writing.

## Character encoding in R code

- **Do not use special characters or accented letters in object, function, column, or file-path
  names in R.** This avoids encoding errors when sharing code between different operating systems
  (Windows/Mac/Linux) or when running the same script on a server with a different *locale*
  setting.
- Always save scripts with explicit **UTF-8** encoding.
- Accented characters can be used in the **narrative text** of an `.Rmd` file (as in this
  document); the restriction applies to code chunks, not to free text (which follows a different
  encoding logic).

## Working language: start in the target language

If the final output is going to a journal in English, the script, comments, object names, and
figures should be written in English **from day one**, not translated at the end as an afterthought.
Concrete reasons, not just style:

- Translating at the end duplicates the work and is a classic source of transcription errors: a
  figure axis translated by hand, a label left half-translated, a comment in the original language
  that no longer matches the (now-translated) name of the object it describes.
- If the dataset or code is eventually deposited in a repository or shared as supplementary
  material, consistent language is part of FAIR **Interoperability**: a script mixing two
  languages is harder for any other group to reuse.
- This doesn't conflict with the previous point: the code (object names, columns, comments) is
  written in the target language of the publication; the narrative prose of an `.Rmd` for internal
  lab use can remain in whatever language the lab normally works in.

## Thinking long-term

All of the practices above — R projects, clear names, backups, version control, documented R
Markdown, consistent language — answer a single question: **could someone else, or myself a year
from now, understand and reproduce this without having to ask?** That is, at bottom, the practical
definition of the R in FAIR (*Reusable*): it's not enough for the analysis to have worked once on
one computer; it has to be able to run again on another computer, with someone else at the
keyboard, and reach the same result.

---

# Common dataset errors and how to prevent them structurally

These are the most common error patterns in datasets built by hand or migrated from Excel, along
with how to eliminate them at the root instead of fixing them case by case:

1. **Never capture a compound field as free text** (e.g., a full scientific name, an address, or
   any field that combines several pieces of information). Always build it by programmatic
   concatenation from atomic columns (`genus`, `specificEpithet`, `taxonRank`, etc., in the case
   of biodiversity data). This eliminates the most frequent errors at the root: inconsistent
   punctuation, missing authorship, misassigned categories, unseparated sub-ranks.
2. **Versioned lookup tables** for any repeated categorical field (administrative geography,
   higher-level taxonomy, experimental treatment type), with a validation `left_join` before every
   export, instead of free-text entry.
3. **A single, global unique identifier** per dataset, not a different one per sub-collection or
   sub-table. If groups need to be distinguished, use a structured prefix (`angio:10040`,
   `pteri:10040`) from the design stage of the database, not patched in afterward.
4. **A pre-submission validation script** (see Section 5) that runs automatically before exporting
   any dataset or before submitting a manuscript, not only when an external reviewer asks for it.
5. **Physical range checks** for every numeric field with physical meaning (elevation,
   coordinates, depth, temperature, any body measurement), with reasonable limits defined in
   advance.
6. **Automatic cross-checking of redundant columns** (for example, a full date versus separate
   year/month/day) at every dataset update, not as a one-off audit years later.
7. **Routine detection of non-printing characters** (non-breaking spaces, curly quotes, em dashes)
   as part of standard quality control, not as a surprise finding from an external reviewer.
8. **Metadata record = code record.** If a manuscript or a README describes a field, that field
   must be generated by the same pipeline that generates the table, so the two never fall out of
   sync.

---

# Automated checking functions

These R functions serve as a quick-audit template for any lab dataset with a tabular structure
(identifier, dates, geography, numeric fields with physical meaning, a compound name/category
field). They are illustrated here with a minimal synthetic dataset structured as biodiversity
occurrence data (Darwin Core standard), but the same logic applies equally to experimental,
field, or any other kind of data: it's just a matter of adjusting the column names.

```{r synthetic-data, echo=FALSE}
# Minimal synthetic dataset, for illustrative purposes only for the checking functions.
occ <- tribble(
  ~occurrenceID, ~catalogNumber, ~collectionCode, ~eventDate, ~year, ~month, ~day,
  ~country, ~continent, ~stateProvince, ~decimalLatitude, ~decimalLongitude,
  ~maximumElevationInMeters, ~scientificName, ~genus, ~specificEpithet, ~taxonRank,

  "PUCV:herb:pteri:10110", "10110", "pteri", "1969-01-01", 1969, 1, 1,
  "Chile", NA_character_, "Los Lagos", -41.833, -73,
  NA_real_, "Hymenophyllum plicatum Kaulf.", "Hymenophyllum", "plicatum", "species",

  "PUCV:herb:angio:5215", "5215", "angio", NA, 1974, NA, NA,
  "Chile", NA_character_, "Valparaíso", -33.4, -71.133,
  40000, "Adiantum thalictroides var. hirsutum (Hook. & Grev.) de la Sota",
  "Adiantum", "chilense", "species",

  "PUCV:herb:angio:6032", NA, "angio", NA, NA, NA, NA,
  NA, NA_character_, NA, NA_real_, NA_real_,
  NA_real_, "Chuquiraga spinosa subsp. rotundifolia (Wedd.) C.Ezcurra",
  "Chuquiraga", "spinosa", "species",

  "PUCV:herb:angio:1806", "1806", "angio", "1958", 1974, NA, NA,
  "Chile", NA_character_, "Valparaíso", -33.617, -70.3,
  NA_real_, "Adesmia boronoides Hook.f.. 1845", "Adesmia", "boronoides", "species",

  "PUCV:herb:angio:9999", "1806", "angio", "1958", 1974, NA, NA,
  "Chile", NA_character_, "Metropolitana", -33.617, -70.3,
  NA_real_, "Adesmia\u00A0boronoides Hook.f. 1845", "Adesmia", "boronoides", "species"
)

kable(occ, caption = "Example synthetic dataset (Darwin Core-type structure)")
```

### 1. Duplicate unique identifiers

```{r check-duplicates, echo=FALSE}
check_duplicates <- function(df, id_col) {
  df %>%
    count(.data[[id_col]], name = "n_records") %>%
    filter(n_records > 1)
}

check_duplicates(occ, "catalogNumber")
```

### 2. Consistency between the full date and year/month/day

```{r check-dates, echo=FALSE}
check_dates <- function(df) {
  df %>%
    mutate(
      year_from_eventDate = suppressWarnings(as.integer(str_extract(eventDate, "^\\d{4}"))),
      matches = is.na(eventDate) | (year_from_eventDate == year)
    ) %>%
    filter(!matches | (is.na(eventDate) & !is.na(year))) %>%
    select(occurrenceID, eventDate, year, month, day)
}

check_dates(occ)
```

### 3. Records with no geographic information at all

```{r check-no-geo, echo=FALSE}
check_no_geography <- function(df) {
  df %>%
    filter(
      is.na(country) & is.na(stateProvince) &
      is.na(decimalLatitude) & is.na(decimalLongitude)
    ) %>%
    select(occurrenceID)
}

check_no_geography(occ)
```

### 4. Values outside a plausible physical range

```{r check-elevation, echo=FALSE}
check_elevation <- function(df, min_m = -430, max_m = 8849) {
  df %>%
    filter(!is.na(maximumElevationInMeters)) %>%
    filter(maximumElevationInMeters < min_m | maximumElevationInMeters > max_m) %>%
    select(occurrenceID, maximumElevationInMeters)
}

check_elevation(occ)
```

### 5. Same coordinate associated with more than one administrative category

```{r check-geo-disagreement, echo=FALSE}
check_geo_disagreement <- function(df) {
  df %>%
    filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
    group_by(decimalLatitude, decimalLongitude) %>%
    summarise(
      n_distinct_provinces = n_distinct(stateProvince, na.rm = TRUE),
      provinces = paste(unique(stateProvince), collapse = " | "),
      .groups = "drop"
    ) %>%
    filter(n_distinct_provinces > 1)
}

check_geo_disagreement(occ)
```

### 6. Invisible characters and duplicated punctuation in text fields

```{r check-characters, echo=FALSE}
check_nbsp <- function(df, col = "scientificName") {
  df %>%
    filter(str_detect(.data[[col]], "\u00A0")) %>%
    mutate(flagged = str_replace_all(.data[[col]], "\u00A0", "{NBSP}")) %>%
    select(occurrenceID, flagged)
}

check_double_period <- function(df, col = "scientificName") {
  df %>%
    filter(str_detect(.data[[col]], "\\.\\.")) %>%
    select(occurrenceID, all_of(col))
}

check_nbsp(occ)
check_double_period(occ)
```

### 7. Compound category not reflected in the rank/level field

```{r check-taxonrank, echo=FALSE}
check_taxonrank_infra <- function(df) {
  patterns <- c("subsp\\.", " var\\.", " f\\.")
  df %>%
    filter(str_detect(scientificName, paste(patterns, collapse = "|"))) %>%
    filter(!taxonRank %in% c("subspecies", "variety", "form")) %>%
    select(occurrenceID, scientificName, taxonRank)
}

check_taxonrank_infra(occ)
```

### 8. Run all checks at once (master function)

```{r check-master, echo=FALSE}
audit_dataset <- function(df) {
  list(
    duplicate_catalogNumber   = check_duplicates(df, "catalogNumber"),
    date_inconsistencies      = check_dates(df),
    no_geography              = check_no_geography(df),
    elevation_out_of_range    = check_elevation(df),
    geo_disagreement          = check_geo_disagreement(df),
    nbsp_in_name              = check_nbsp(df),
    double_period_in_name     = check_double_period(df),
    bad_taxonrank_infra       = check_taxonrank_infra(df)
  )
}

result <- audit_dataset(occ)
result
```

> **To use on real lab data:** save these functions in `R/audit_functions.R`, load them with
> `source("R/audit_functions.R")`, adapt the column names to your own dataset, and run
> `audit_dataset(my_dataset)` as the last step of `scripts/02_clean.R`, before exporting any
> dataset or submitting a manuscript for review.

---

# Pre-submission checklist for lab datasets and manuscripts

```{r checklist, echo=FALSE}
checklist <- tribble(
  ~step, ~description,
  1, "The original raw file is saved, intact, and dated, with no manual editing.",
  2, "At least one backup of the raw file exists off the everyday working machine.",
  3, "The entire cleaning pipeline is in version-controlled code (Git), not manual Excel edits.",
  4, "Compound fields (names, addresses, combined categories) are generated by programmatic concatenation, not free typing.",
  5, "audit_dataset() (or equivalent) was run, and no findings remain unresolved or undocumented.",
  6, "Redundant fields (e.g., full date vs. year/month/day) are 100% consistent with each other.",
  7, "There are no duplicate identifiers within the same scope.",
  8, "All numeric fields with physical meaning fall within plausible ranges.",
  9, "No invisible characters or punctuation errors remain in text fields.",
  10, "The metadata described in the manuscript or README exactly match the exported columns.",
  11, "The repository README documents the cleaning process and the decisions made."
)

kable(checklist, col.names = c("#", "Check")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```

---

# Illustrative case: technical audit of a biodiversity dataset

To avoid staying at the level of abstract principles, it's worth looking at a real example of what
happens when these practices are not followed from the start: the technical report that Pensoft
(the publisher of *PhytoKeys*, *ZooKeys*, and *Biodiversity Data Journal*) commissioned from Dr.
Robert Mesibov on the Herbarium dataset of the Pontificia Universidad Católica de Valparaíso
(PUCV), downloaded from GBIF on 2022-07-13. Pensoft commissions this kind of technical evaluation
before sending a "data paper" (an article whose purpose is to describe a dataset, not to present
research results) out for peer review. The reviewer himself is explicit about the scope of the
check:

> *"Pensoft does not check the details of the content of a dataset, for example whether the
> correct author is given for a scientific name, or whether the correct latitude/longitude is
> given for a locality."*

In other words: it's a check of **structural consistency**, not of the scientific accuracy of the
content.

```{r errors-table, echo=FALSE}
errors_table <- tribble(
  ~category, ~finding, ~severity,
  "Schema/metadata", "Field described in the paper but absent from the table", "High",
  "Invisible characters", "Non-breaking spaces (NBSP) within the scientific name", "Medium",
  "Identifiers", "Duplicate identifier across sub-collections", "High",
  "Dates", "Full date, year, month, and day inconsistent with each other", "High",
  "Geography", "Records with no locality information at all", "High",
  "Geography", "Same coordinate assigned to a different administrative region", "High",
  "Impossible values", "Elevation greater than the maximum plausible terrestrial height", "High",
  "Higher taxonomy", "Family name used as if it were an order", "Medium",
  "Taxonomic rank", "Infraspecific rank not reflected in the corresponding field", "High",
  "Scientific name", "Double period instead of one; missing authorship", "Medium"
)

kable(errors_table, caption = "Sample of the error categories identified in the PUCV dataset (condensed version)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

Nearly all of these problems share a common origin, more important than any single error:
manual capture of fields that should have been built by programmatic concatenation, copy-pasting
from Word or PDF (which introduces invisible characters and duplicated punctuation), the absence
of controlled vocabularies for categorical fields, and a lack of range validation for numeric
fields. None of these origins is exclusive to herbarium data: they are exactly the same patterns
that motivate Sections 4 and 5 of this document.

It's also worth clarifying what this type of audit does **not** cover, so as not to over-trust a
dataset that passed it: it does not verify the scientific correctness of the content (authorship
of a name, the actual coordinates of the collection site), it does not evaluate the taxonomic
identity of the specimens, it does not audit the prior digitization process, and it is a snapshot
in time, not a guarantee that the same pipeline won't reproduce the same errors at the next update.
Passing a technical audit is a necessary but not sufficient condition for data quality: it
complements, but does not replace, content review by a specialist.

---

# Some useful links

- Darwin Core Standard: <https://dwc.tdwg.org/>
- OpenRefine: <https://openrefine.org/>
- Notepad++: <https://notepad-plus-plus.org/>
- R Projects documentation: <https://r4ds.hadley.nz/workflow-scripts>
- GBIF Data Quality Requirements: <https://www.gbif.org/data-quality-requirements-occurrences>
